---
title: "Coexpression Network Analysis for minfi curated methylation data (unsigned)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file is

# Clear R console screen output
cat("\014")  

# Load required libraries
library(synapseClient)
library(data.table)
library(dplyr)
library(knitr)
library(stringr)
library(RColorBrewer)
library(ggplot2)
library(gplots)
library(limma)
library(edgeR)
library(ctv)
library(psych)
library(reshape2)
library(vcd)
library(erer)
library(fpc)
library(WGCNA)
library(Rclusterpp)
library(circlize)
library(knit2synapse)
library(vbsr)

## Needs the dev branch
library(rGithubClient)

## Requires ggdendro
devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")

# source('/mnt/Github/knit2synapse-1/R/knitFile2SynapseFolder.R')
# knitToFolderEntity(file = "./CoexppNet_methyl_TechCovar.Rmd", 
#                    entityName = 'Coexpression Network Analysis methyl All Technical Covariates Model Unsigned', 
#                    parentId = 'syn4231339', 
#                    overwrite=F)

synapseLogin()

# source utility files from ../R/lib folder
file.sources = list.files('../R/lib',pattern="*.R")
file.sources = sapply(file.sources,function(x){return(paste('../R/lib',x,sep='/'))})
tmp = sapply(file.sources,source,.GlobalEnv)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE)
```
### Download data
```{r setup, include=FALSE}
# Input Parameters
EXP_ID = 'syn2233188'
METADATA_ID = 'syn3156828'

parentId = 'syn4231339'

# Specify factor and continuous covarites pool (adjusted covariates will only be the subset of these covariates)
FactorCovariates = c("Diffname_short", "BeadChip", "Row", "Column", "Cell_Line_Type", 
                     "Tissue_of_Origin", "Reprogramming_Gene_Combination", "Culture_Conditions",  
                     "Other_Conditions_During_Reprogramming", "Donor_Life_Stage", "Gender", 
                     "Originating_Lab", "Donor_ID", "Cell_Type_of_Origin_Level2",
                     "Reprogramming_Vector_Type" )
ContCovariates = c("PassageAtThaw", "PassageAtDNAHarvest")
```
Obtain count matrix and metadata from synapse
```{r getdata, cache=TRUE, include=FALSE}
# Get count matrix
EXP_OBJ = synGet(EXP_ID)
ALL_USED_IDs = EXP_OBJ$properties$id
EXP = fread(getFileLocation(EXP_OBJ), data.table = F)
row.names(EXP) = EXP[,1]
EXP = dplyr::select(EXP,-(ProbeID))

# Get metadata
METADATA_OBJ = synTableQuery(paste('SELECT * FROM',METADATA_ID,sep=' '), loadResult=F)
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_OBJ@schema
METADATA = fread(METADATA_OBJ@filePath, data.table=F)
```
Synapse id of count matrix used for the analysis is `r paste(EXP_ID, EXP_OBJ$properties$versionNumber,sep='.')` and the synapse id of meta data table used for the analysis is `r METADATA_ID` 

Factor covariates considered for analysis are `r paste(gsub('_','\\\\_',FactorCovariates),collapse=',')`, and continuous covariates considered for the analysis are `r paste(gsub('_','\\\\_',ContCovariates),collapse=',')`

Preprocess counts matrix and metadata
```{r preprocessing, include=FALSE}
# Preprocess metadata
METADATA[METADATA == 'N/A'] = NA

# Replace all special characters with blank
myFix <- function(x) str_replace_all(x, '[^[:alnum:]]', '')
METADATA <- METADATA %>%
  dplyr::mutate_each(funs(myFix), -UID, -C4_Cell_Line_ID, -biologicalSampleName) # fix them but don't touch some columns

# Set rownames
rownames(METADATA) = METADATA$UID
```
### Preprocess data
* Remove samples with no Diffname short
* Remove somatic samples and samples with no Cell Type
* Remove samples with abnornal or no C4 Karyotype Result
```{r filtering, echo=TRUE}
#### Pre processing mRNA expression counts and metadata ####
metadata_filtered <- 
  METADATA %>%
  filter(Diffname_short != "") %>%
  filter(UID != "H9P50") %>%
  filter(UID %in% colnames(EXP)) %>%
  filter(Cell_Type == "PSC") %>%
  filter(C4_Karyotype_Result != "abnormal")

REMOVED_UID <- setdiff(colnames(EXP), metadata_filtered$UID)
METADATA <- METADATA[metadata_filtered$UID,]
EXP <- EXP[, METADATA$UID]
```
The following `r length(REMOVED_UID)` samples were removed:

`r paste(gsub('_','\\\\_',REMOVED_UID),collapse=',')` 

### Covariate clustering
Determine relationship between covariates
```{r covariates.clustering}
# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates)]

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
```
### Probe filtering
Retain probes with beta greater than 0.5 in atleast one sample and less than 0.5 in atleast one other sample
```{r gene.filtering, cache=TRUE, fig.width=8, fig.height=8, echo=TRUE}
exp.variance <- apply(as.matrix(EXP),1,var)
ind.remove1 <- exp.variance < sort(exp.variance)[round(0.25*length(exp.variance))]
ind.remove2 <- (rowSums(EXP <= 0.25)/dim(EXP)[2] == 1) | (rowSums(EXP >= 0.75)/dim(EXP)[2] == 1)

EXP <- EXP[!ind.remove1 & !ind.remove2,]
```
`r dim(EXP)[1]` number of probes and `r dim(EXP)[2]` number of samples are considered for the analysis

### Clustering samples
Clustering of initial data (with NULL design)
```{r decompse.normalise.data, cache=TRUE, fig.height=8, fig.width=8}
# Find principal components of expression to plot
PC <- prcomp(EXP)

# Plot first 2 PCs
plotdata <- data.frame(UID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- merge(plotdata, METADATA, by="UID")
plotdata <- mutate(plotdata, labels = '');#ifelse(Diffname_short == '',as.character(UID),'')) 

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color = Cell_Line_Type, shape=Diffname_short, size=Gender))
p <- p + theme_bw() + scale_size_manual(values = c(5, 2)) + scale_shape_manual(values = 1:7) + theme(legend.position="top") 
p <- p + geom_text(aes(label= labels), size=4, hjust=0)
p
```
### Normalisation (with technical covariates only)
Initial normalisation usign voom (residualised technical covariates with donor id as random variable)
```{r technical.voom.normalisation, fig.height=14, fig.width=20}
postAdjustCovars = c('Diffname_short', 'Gender', 'Culture_Conditions', 'BeadChip', 'Row', 'Donor_Life_Stage')
  
# Post adjusted design matrix
DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
    
# Estimate correlation of random effects
correlation <- duplicateCorrelation(EXP, design=DM1$design, block=COVARIATES$Donor_ID)
  
# Fit linear model
FIT = lmFit(EXP, design=DM1$design, block=COVARIATES$Donor_ID, correlation = correlation$cor)
  
# Residuals after normalisation
EXPR = residuals.MArrayLM(FIT,EXP)
varsToAddBackIn = c("Diffname_shortDE", "Diffname_shortEB", "Diffname_shortECTO", "Diffname_shortMESO5", "Diffname_shortSC", "Genderfemale", "Culture_ConditionsCreexcised", "Culture_ConditionsDMEMF12KOSR01mMBME01mMNEAA01mMLgluatime20ngmlbFGF", "Culture_ConditionsE8", "Culture_ConditionsFeederandKOSRcultureconditions", "Culture_ConditionsStandardTESR1MatrigelorKOSRMEFs", "Culture_ConditionsStromalnonprimed", "Culture_ConditionsStromalprimed", "Culture_ConditionsiPSCsweregeneratedandintiallymaintainedusingMEFKSRculturesAftertransfertoWiCelliPSCsexpandedinFeederIndependentTeSR1Medium", "Donor_Life_Stageadult", "Donor_Life_Stageblastocyst", "Donor_Life_Stageembryo", "Donor_Life_Stagefetal")
EXPR = EXPR + (FIT$coefficients[,varsToAddBackIn] %*% t(DM1$design[,varsToAddBackIn]))

# Design matrix with all covariates
DESIGN = getDesignMatrix(COVARIATES, Intercept = F)
DESIGN = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
colnames(DESIGN) = str_sub(colnames(DESIGN), 1, 30)

# Correlation between covariates and expression residuals
residualCovaraitesCorrelation = runPCAandPlotCorrelations(EXPR, DESIGN, dataName = 'residual covariates relation')
print(residualCovaraitesCorrelation$PC_res[[2]])
```
Sample clustering to check for outliers
```{r sample.clustering, fig.height=15, fig.width=20}
sampleTrees = flashClust::hclust(dist(t(EXPR)), method = "average")
tree = cutreeStatic(sampleTrees, cutHeight = 200, minSize=3)

plotDendroAndColors(sampleTrees, 
                    labels2colors(cbind(tree,COVARIATES)),
                    groupLabels = c('Cluster',colnames(COVARIATES)))
collectGarbage()
```
Sample pca to check for outliers
```{r residual.clustering, fig.height=10, fig.width=10}
# Find principal components of expression to plot
PC <- prcomp(EXPR)

# Determine number of clusters automatically using pamk
pam.cluster <- pamk(t(EXPR), krange=2:10)

# Plot first 2 PCs
plotdata <- data.frame(UID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2], 
                       cluster=pam.cluster$pamobject$clustering)

plotdata <- merge(plotdata, rownameToFirstColumn(COVARIATES, "UID"), by="UID")
plotdata <- mutate(plotdata, labels = '');#ifelse(Diffname_short == '',as.character(UID),'')) 

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=factor(cluster), shape=Diffname_short, size=Gender))
p <- p + theme_bw() + scale_size_manual(values = c(4, 2)) + scale_shape_manual(values=1:7) + theme(legend.position="top") 
p <- p + geom_text(aes(label= labels), size=4, hjust=0)
p
```
### Coexpression networks
```{r wgcna.power, fig.height=8, fig.width=8}
# Allow multi-threading within WGCNA. 
# Caution: skip this line if you run RStudio or other third-party R environments.
enableWGCNAThreads()

# Choose a set of soft-thresholding powers
powers = c(seq(2,10,by=1), seq(12,30, by=2));

# Call the network topology analysis function
EXPR = t(EXPR)
powerTables = pickSoftThreshold(EXPR, 
                                powerVector=powers,
                                corFnc = 'bicor',
                                networkType = 'unsigned',
                                blockSize = 10000,
                                verbose = 2)[[2]]
collectGarbage();

# Plot the results
colors = c("black", "red")

# Will plot these columns of the returned scale free analysis tables
plotCols = c(2,5,6,7)
colNames = c("Scale Free Topology Model Fit", "Mean connectivity", "Median connectivity", "Max connectivity");

# Plot the quantities in the chosen columns vs. the soft thresholding power
p <- list()
for (i in 1:length(plotCols)){
  name = colnames(powerTables)[plotCols[i]]
  p[[name]] <- ggplot(powerTables, aes_string(x = 'Power', y = name)) + geom_point()
  p[[name]] <- p[[name]] + geom_text(aes(label= Power), size=4, hjust=1, vjust = -1)
  p[[name]] <- p[[name]] + ggtitle(colNames[i])
}

multiplot(plotlist=p, cols=2)
softPower = 7
collectGarbage()
save(list=c('powerTables'), file = 'CoexppNet_methyl_technicalCovariates_unsigned.RData')
```
Beta is chosen to be `r softPower`
```{r wgcna.adj, fig.width=15, fig.height=10}
# We set the minimum module size relatively high
minModuleSize = 30;

# Get blockwise modules 
MOD = blockwiseModules(EXPR,
                       maxBlocksize = 10000,
                       corType = "bicor",
                       power = softPower,
                       networkType = "unsigned",
                       deepSplit = 3,
                       minModuleSize = minModuleSize,
                       reassignThershold = 1e-20,
                       verbose = 4)
collectGarbage()

# Module identification using dynamic tree cut
unmergedNetModules.colors = MOD$colors
collectGarbage()

save(list=c('powerTables', 'MOD', 'unmergedNetModules.colors'), 
     file = 'CoexppNet_methyl_technicalCovariates_unsigned.RData')
```
Dendrogram tree of gene clustering
```{r dendro.modules, fig.height=12, fig.width=20}
for (i in 1:length(MOD$blockGenes)){
  plotDendroAndColors(MOD$dendrograms[[i]], unmergedNetModules.colors[MOD$blockGenes[[i]]], 
                      "Dynamic Tree Cut", dendroLabels = FALSE, hang = 0.03,
                      addGuide = TRUE, guideHang = 0.05)
  }
```
### Module analysis
```{r module.analysis, fig.height=5, fig.width=10}
# Merge modules with similar expression profiles
# Calculate module eigengenes
unmergedMEs = moduleEigengenes(EXPR, 
                               colors = unmergedNetModules.colors,
                               nPC = 10)

# Calculate dissimilarity between module eigen genes
MEDiss = 1 - abs(cor(unmergedMEs$eigengenes, use = "p"))

# Cluster module eigen genes
METree = hclust(as.dist(MEDiss), method = "average");
collectGarbage()

# Plot the results
par(mfrow = c(1,1))
plot(METree, main = "Clustering of module eigengenes",
     xlab = "", sub = "")
abline(h=0.15, col = "red")

# Merge modules
merge = mergeCloseModules(fixDataStructure(EXPR), unmergedNetModules.colors, cutHeight = 0.15, verbose = 3)

# Numeric module labels
mergedNetModules = merge$colors

# Convert labels to colors
mergedNetModules.colors = labels2colors(mergedNetModules)
mergedNetModules.colors = matchLabels(unmergedNetModules.colors, mergedNetModules.colors)

# Eigengenes of the new merged modules:
mergedMEs = merge$newMEs
MEs = orderMEs(mergedMEs[[1]]$data)

tmp = corr.test(MEs)
tmp1 = tmp$r
tmp1[tmp$pval >= 0.05] = 0
tmp1 = melt(tmp1)

# Plot correlation between eigen genes
writeLines('Correlation between eigen genes')
p = ggplot(tmp1, aes(x=X1,y=X2)) + geom_tile(aes(fill = value), colour = "white") + 
  scale_fill_gradient2(low = "green", high = "red", limits = c(-1,1))
p = p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
p

# Plot dendrogram and module colors
writeLines('Module dendrogram and color assignment')
for (i in 1:length(MOD$dendrograms)){
  plotDendroAndColors(MOD$dendrograms[[i]],
                      cbind(unmergedNetModules.colors[MOD$blocks == i], mergedNetModules.colors[MOD$blocks == i]),
                      c("Unmerged", "Merged"),
                      dendroLabels = FALSE, hang = 0.03,
                      addGuide = TRUE, guideHang = 0.05)
  }
save(list=c('powerTables', 'MOD', 'unmergedNetModules.colors',
            'mergedNetModules.colors', 'mergedMEs'), 
     file = 'CoexppNet_methyl_technicalCovariates_unsigned.RData')
```
### Compare eigen genes and covariates
```{r compare.covars, fig.height=10, fig.width=15}
# Set up variables to contain the module-trait correlations
DESIGN = getDesignMatrix(COVARIATES[,FactorCovariates], Intercept = F)$design
DESIGN[is.na(DESIGN)] = 0

DESIGN.SPLIT = lapply(FactorCovariates, function(x, DESIGN){ ind = grep(x, colnames(DESIGN)); return(DESIGN[,ind])}, DESIGN)
names(DESIGN.SPLIT) = FactorCovariates

# Also obtain comparisons
ALL.COMP = mapply(function(x, y){
  CONT.NAMES <- colnames(x)
  CONT.NAMES <- combn(CONT.NAMES,2)
  CONT.NAMES <- apply(CONT.NAMES,2,paste,collapse='-')
  
  CONT <- makeContrasts(contrasts=CONT.NAMES,levels=colnames(x))
  colnames(CONT) = gsub(y,'',colnames(CONT))
  
  ALL.CONT = x %*% CONT
  
  colnames(x) = gsub(y, '', colnames(x))
  
  return(ALL.CONT)
  }, DESIGN.SPLIT, FactorCovariates)
ALL.COMP = do.call(cbind, ALL.COMP)
ALL.COMP = ALL.COMP[,!duplicated(colnames(ALL.COMP))]
```
Correlation between module eigen genes and molecular traits
```{r calculate.corr, fig.height=15, fig.width=10}
# Calculate the correlations
g = corr.test(mergedMEs[[1]]$data, ALL.COMP, adjust = 'bonferroni')

moduleTraitCor = g$r
moduleTraitPvalue = g$p
moduleTraitCor[moduleTraitPvalue > 0.05] = 0

moduleTraitCor = moduleTraitCor[, colSums(moduleTraitCor) != 0]
colnames(moduleTraitCor) = str_sub(colnames(moduleTraitCor),1,25)

tmp = reshape2::melt(moduleTraitCor)
setnames(tmp, c('X1','X2'), c('MEs','Trait'))
p = ggplot(tmp, aes(x=MEs,y=Trait)) + geom_tile(aes(fill = value), colour = "white") + 
  scale_fill_gradient2(low = "green", high = "red", limits = c(-1,1))
p = p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
p

# Heatmap(moduleTraitCor, 
#         column_title = 'Covariates', 
#         row_title = 'Module Eigengene', 
#         name = 'Correlation',
#         col = colorRamp2(c(-1,0,1), c("green", "white", "red")))

save(list=c('powerTables', 'MOD', 'unmergedNetModules.colors',
            'mergedNetModules.colors', 'mergedMEs','g','ALL.COMP'), 
     file = 'CoexppNet_methyl_technicalCovariates_unsigned.RData')
```
vbsr between MEs and traits
```{r vbsr, fig.height=15, fig.width=10}
associationStats = sapply(MEs, function(x, ALL.COMP){
  tmp = vbsr(x,ALL.COMP)$z
  names(tmp) = colnames(ALL.COMP)
  return(tmp)
}, ALL.COMP)
rownames(associationStats) = str_sub(rownames(associationStats), 1, 25)

associationStats = associationStats[rowSums(abs(associationStats) >= 5) != 0,colSums(abs(associationStats) >= 5) != 0]

tmp = reshape2::melt(associationStats)
setnames(tmp, c('X1','X2','value'), c('Trait', 'MEs','Zscores'))
p = ggplot(tmp, aes(x=MEs,y=Trait)) + geom_tile(aes(fill = Zscores), colour = "white") + 
  scale_fill_gradient2(low = "green", high = "red", limits = c(-70,70))
p = p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
p

save(list=c('powerTables', 'MOD', 'unmergedNetModules.colors',
            'mergedNetModules.colors', 'mergedMEs','g','ALL.COMP','associationStats'), 
     file = 'CoexppNet_methyl_technicalCovariates_unsigned.RData')
```

### Synapse Store
```{r synapse.store, cache=FALSE, include=FALSE}
parentId = 'syn3256504'

activityName='Coexpression Network Analysis of methyl data with technical covariates model'
  
thisFileName <- 'CoexppNet_methyl_TechCovar.Rmd'
  
# Github link
thisRepo <- getRepo(repository = "th1vairam/pcbc_c4_analysis", 
                    ref="branch", 
                    refName='methylNet')

thisFile <- getPermlink(repository = thisRepo,
                        repositoryPath=paste0('code/Rmd/', thisFileName))
    
# Create folder to tore the results and populate analysis wiki
CODE <- Folder(name = 'Coexpression Network Analysis methyl All Technical Covariates Model Unsigned',parentId = parentId)
CODE <- synStore(CODE)
      
# Store Results
COEXPP_OBJ <- File('./CoexppNet_methyl_technicalCovariates_unsigned.RData', name = 'Coexpression Networks', parentId = CODE$properties$id)
COEXPP_OBJ <- synStore(COEXPP_OBJ, used = ALL_USED_IDs, activityName = activityName, executed=thisFile)

# Store Results
MOD = data.frame(GeneNames = colnames(EXPR) , merged = mergedNetModules.colors, unmerged = unmergedNetModules.colors)
write.table(MOD, file='modules.tsv', sep='\t', quote=F, row.names=F)
MOD_OBJ <- File('modules.tsv', name = 'Coexpression Network Modules', parentId = CODE$properties$id)
MOD_OBJ <- synStore(MOD_OBJ, used = ALL_USED_IDs, activityName = activityName, executed=thisFile)
```
### Source Code
[Source R Markdown](`r thisFile`)