---
title: "miRNA descriptive analysis"
author: "Lorena Pantano"
date: "02/09/2015"
output: html_document
---

```{r libs, warning=FALSE, message=FALSE}
library(tidyr)
library(dplyr)
library(knitr)
library(synapseClient)
library(DESeq2)
library(vcd)
source("../R/lib/qc.R")
library(corrplot)

synapseLogin()

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=TRUE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE)
```

# Load Data

## Get metadata and count matrix 

```{r get-id}
Query <- "select id,UID,name,fileType,dataType from file where id=='syn2247832' AND  projectId=='syn1773109'"
res <- synQuery(Query)
mat_file = synGet(res$file.id)

Query <- "select id,UID,name,fileType,dataType from file where id=='syn2278179' AND  projectId=='syn1773109'"
res <- synQuery(Query)
meta_file = synGet(res$file.id)
```


## Clean and sync tables

Just make sure the tables are sync, same order for samples

```{r getcountmat}
meta = read.table(meta_file@filePath,header=1,row.names=1,sep="\t")
row.names(meta) = gsub("-",".",row.names(meta))
mat = read.table(mat_file@filePath,header=T,row.names=1,sep="\t")

meta_clean = meta[colnames(mat), ]
mat[is.na(mat)] = 0
```

# Results

## Correlation of metadata variables

I looked for a way to get a correlation p-value and a rho value for categorical
variables. I found that, that is quite similar, and could give us an idea.

```{r cor}
cols = c("originating.lab",
         "Diffname.short","Gender","Tissue.of.Origin",
         "Cell.Type.of.Origin","Cell.Line.Type","Reprogramming.Gene.Combination")
cols_format = sub("[::.::]","\n",cols) # reduce space of labels in heapmap

comp = expand.grid(cols,cols)
c = apply(comp,1,corr_two_nominal_vars,meta_clean)
res = cbind(comp,do.call(rbind,c))
names(res)[3:4] = c("cor","pvalue")

cor_mat = tidyr::spread(res[,c(1,2,3)],"Var1","cor")
cor_mat = cor_mat[,2:ncol(cor_mat)]
cor_mat = matrix(unlist(cor_mat),nrow=length(cols))
row.names(cor_mat) = cols_format
colnames(cor_mat) = cols_format

pval_mat = tidyr::spread(res[,c(1,2,4)],"Var1","pvalue")
pval_mat = pval_mat[,2:ncol(pval_mat)]
pval_mat = matrix(unlist(pval_mat), nrow=length(cols))
row.names(pval_mat) = cols_format
colnames(pval_mat) = cols_format

cor_mat[is.na(cor_mat)] = 0
corrplot(cor_mat, p.mat = pval_mat, method = 'number', insig = "blank",  
         is.corr=TRUE, sig.level = 0.05,tl.cex = 0.7, cl.cex=0.7)
```

## PCA

There are some values that correlates, with PCA should be easy to see if samples
have some bias to the previous variables. If there is any, this should be taking
into account during DE.

I used DESeq2 because for miRNA, I always found to normalized better, since there is quite variation in the top expressed miRNA, and many times bias to library preparation.

If we want to look for a general UP/DOWN regulation, this is not valid.

```{r deseq2, cache=TRUE}
dds = DESeqDataSetFromMatrix(mat[rowMeans(mat)>5,], colData = meta_clean, design =~1)
rlogMat = rlog(dds, blind= TRUE)
```

I tried many different variables, to make sure there wasn't a bias.

I expected some bias due to Donor.life.Stage, although blastocyst are mixed
with adult and fetal samples. 

It seems are separated by differentiation stage, but don't know if it makes sense.

```{r plotpca}
DESeq2::plotPCA(rlogMat, intgroup = "Gender", ntop=200)
DESeq2::plotPCA(rlogMat, intgroup = "Cell.Type.of.Origin", ntop=200)
DESeq2::plotPCA(rlogMat, intgroup = "Tissue.of.Origin", ntop=200)
DESeq2::plotPCA(rlogMat, intgroup = "originating.lab", ntop=200)
DESeq2::plotPCA(rlogMat, intgroup = "Reprogramming.Gene.Combination", ntop=200)
DESeq2::plotPCA(rlogMat, intgroup = "Cell.Line.Type", ntop=200)
DESeq2::plotPCA(rlogMat, intgroup = "CellDiffState", ntop=200)
DESeq2::plotPCA(rlogMat, intgroup = "GroupLevel1.DifferentiationState", ntop=200)
DESeq2::plotPCA(rlogMat, intgroup = "GroupLevel3.DifferentiationState", ntop=200)


DESeq2::plotPCA(rlogMat, intgroup = "Donor.Life.Stage", ntop=200)
DESeq2::plotPCA(rlogMat, intgroup = "GroupLevel2.DifferentiationState", ntop=200)
```


