---
title: "Covariate analysis of minfi curated methylation data with mixed effect modeling performed at individual differentiation stages"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2syn, eval = F, echo = T}
library(synapseClient)
library(knitr)
library(githubr)

synapseLogin()

knit2synapse::knitToFolderEntity(file = "./CovariateAnalysis_methylation_mixedEffects_EachDiffState_withECTOScores.Rmd",
                                 parentId = 'syn4231339',
                                 entityName = 'Covariate Analysis Methylation Mixed Effects EachDiffState withECTOScores')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file is

# Clear R console screen output
cat("\014")  

# Load required libraries
library(synapseClient)
library(knitr)
library(githubr)

library(CovariateAnalysis)
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library('RColorBrewer')
library('ggplot2')
library('gplots')
library('limma')
library('edgeR')
library('ctv')
library('psych')
library('reshape2')
library('vcd')
library('erer')
library('fpc')
library('stringr')
library('minfi')

## Requires ggdendro
devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")

synapseLogin()

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE)
```
### Download data
```{r setup, include=FALSE, cache = FALSE}
# Input Parameters
EXP_ID = 'syn2233188'
METADATA_ID = 'syn3156828'

parentId = 'syn4231339'

# Specify factor and continuous covarites pool (adjusted covariates will only be the subset of these covariates)
FactorCovariates = c("Diffname_short", "BeadChip", "Row", "Column", "Cell_Line_Type", 
                     "Tissue_of_Origin", "Reprogramming_Gene_Combination", "Culture_Conditions",  
                     "Other_Conditions_During_Reprogramming", "Donor_Life_Stage", "Gender", 
                     "Originating_Lab", "Donor_ID", "Cell_Type_of_Origin_Level2",
                     "Reprogramming_Vector_Type" )
ContCovariates = c("PassageAtThaw", "PassageAtDNAHarvest", "ecto.scores")
```
Factor covariates considered for analysis are `r paste(gsub('_','\\\\_',FactorCovariates),collapse=',')`, and continuous covariates considered for the analysis are `r paste(gsub('_','\\\\_',ContCovariates),collapse=',')`

Obtain count matrix and metadata from synapse
```{r getdata, cache=FALSE, include=FALSE}
# Get count matrix
EXP_OBJ = synGet(EXP_ID)
ALL_USED_IDs = EXP_OBJ$properties$id
EXP = fread(getFileLocation(EXP_OBJ), data.table = F)
row.names(EXP) = EXP[,1]
EXP = dplyr::select(EXP,-(ProbeID))

# Get metadata
METADATA_OBJ = synTableQuery(paste('SELECT * FROM',METADATA_ID,sep=' '))
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_OBJ@schema
METADATA = METADATA_OBJ@values

# Get cell line specific meso-ecto scores and add 1to metadata
cibersort.scores = read.csv(synGet('syn7254021')@filePath, header = T) %>%
  dplyr::mutate(ecto.scores = ECTO)  %>%
  dplyr::rename(UID = Input.Sample) %>%
  tidyr::separate(UID, c('C4_Cell_Line_ID','a','b','c'), sep = '\\.') %>%
  dplyr::mutate(C4_Cell_Line_ID = gsub('EB','',C4_Cell_Line_ID)) %>%
  group_by(C4_Cell_Line_ID) %>%
  summarise(ecto.scores = mean(ecto.scores, na.rm=T))
ALL_USED_IDs <- c(ALL_USED_IDs, 'syn7254021')

METADATA = left_join(METADATA, cibersort.scores)
```
Synapse id of count matrix used for the analysis is `r paste(EXP_OBJ$properties$id,EXP_OBJ$properties$versionNumber,sep='.')` and the synapse id of meta data table used for the analysis is `r METADATA_ID` 

Preprocess counts matrix and metadata
```{r preprocessing, include=FALSE, cache = FALSE}
# Preprocess metadata
METADATA[METADATA == 'N/A'] = NA

# Replace all special characters with blank
myFix <- function(x) str_replace_all(x, '[^[:alnum:]]', '')
METADATA <- METADATA %>%
  dplyr::mutate_each(funs(myFix), -UID, -C4_Cell_Line_ID, -biologicalSampleName, -ecto.scores) # fix them but don't touch some columns

# Set rownames
rownames(METADATA) = METADATA$UID
```
### Preprocess data
* Remove samples with no Diffname short
* Remove somatic samples and samples with no Cell Type
* Remove samples with abnornal or no C4 Karyotype Result
```{r filtering, echo=TRUE, cache = FALSE}
#### Pre processing mRNA expression counts and metadata ####
metadata_filtered <- 
  METADATA %>%
  filter(Diffname_short != "") %>%
  filter(UID != "H9P50") %>%
  filter(UID %in% colnames(EXP)) %>%
  filter(Cell_Type == "PSC") %>%
  filter(C4_Karyotype_Result != "abnormal" | !is.na(C4_Karyotype_Result)) %>%
  filter(!is.na(ecto.scores))

REMOVED_UID <- setdiff(colnames(EXP), metadata_filtered$UID)
METADATA <- METADATA[metadata_filtered$UID,]
EXP <- EXP[, METADATA$UID]
```
The following `r length(REMOVED_UID)` samples were removed:
  
  `r paste(gsub('_','\\\\_',REMOVED_UID),collapse=', ')` 
### Covariate clustering
Determine relationship between covariates
```{r covariates.clustering,cache=FALSE}
# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates)]

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)

# Split covariates 
COVARIATES = split(COVARIATES,COVARIATES$Diffname_short)
```
Covariate correlation
```{r covariates.correlation, fig.width=10, fig.height=10, cache = FALSE}
covariateCorrelation <- function(COVARIATES){  
  Diffname = unique(COVARIATES$Diffname_short)
  writeLines('------------------------------------------------------------------------------')
  writeLines(paste('Differentiation state:',as.character(Diffname)))
  writeLines('------------------------------------------------------------------------------')
  
  COVARIATES = droplevels(COVARIATES)
  COVARIATES = COVARIATES[,c(which(sapply(COVARIATES, function(x) length(levels(x))) > 1),
                             which(colnames(COVARIATES) %in% ContCovariates))]  
  # Correlation between covariates
  COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.1)  
  print(ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.3, row.width=0.15))
  
  return(list(COVARIATES.CORRELATION = COVARIATES.CORRELATION))
}

COVARIATES.CORRELATION = lapply(COVARIATES, covariateCorrelation)
```
### Probe filtering
Retain probes with beta greater than 0.5 in atleast one sample and less than 0.5 in atleast one other sample
```{r gene.filtering, cache=TRUE, fig.width=8, fig.height=8, echo=TRUE, cache = FALSE}
splitData <- function(cols,EXP,METADATA){
  Diffname = unique(METADATA[cols,'Diffname_short'])
  writeLines('------------------------------------------------------------------------------');
  writeLines(paste('Differentiation state:',as.character(Diffname)));
  writeLines('------------------------------------------------------------------------------');
  
  ind.remove <- (rowSums(EXP[,cols] <= 0.2)/dim(EXP[,cols])[2] == 1) | (rowSums(EXP[,cols] >= 0.7)/dim(EXP[,cols])[2] == 1)
  
  EXP <- EXP[!ind.remove,cols]  
  
  writeLines(paste(dim(EXP)[1],'number of probes and',dim(EXP)[2],'number of samples are considered for the analysis'));
  return(EXP)
}

PROC.EXP <- tapply(colnames(EXP),
                   factor(METADATA$Diffname_short),
                   splitData,
                   EXP,
                   METADATA)
```

Clustering of initial normalised data (with NULL design)
```{r decompse.normalise.data, include=TRUE, cache=TRUE, fig.height=8, fig.width=8, cache = FALSE}
plotPCAMethylExpression <- function(COVARIATES, PROC.EXP){ 
  writeLines('------------------------------------------------------------------------------')
  diffstate = unique(COVARIATES$Diffname_short)
  writeLines(paste('Differentiation state:',as.character(diffstate)))
  writeLines('------------------------------------------------------------------------------')
  
  # Drop unilevel covariates
  COVARIATES = droplevels(COVARIATES)
  COVARIATES = COVARIATES[,c(which(sapply(COVARIATES, function(x) length(levels(x))) > 1),
                             which(colnames(COVARIATES) %in% ContCovariates))]    
  
  # Find principal components of expression to plot
  PC <- prcomp(PROC.EXP[[diffstate]])
  
  # Plot first 2 PCs
  plotdata <- data.frame(UID = rownames(PC$rotation), 
                         PC1 = PC$rotation[,1], 
                         PC2 = PC$rotation[,2])
  
  plotdata <- merge(plotdata, METADATA[colnames(PROC.EXP[[diffstate]]),], by="UID")
  plotdata <- mutate(plotdata, labels = '');#ifelse(Diffname_short == '',as.character(UID),'')) 
  
  p <- ggplot(plotdata, aes(x=PC1, y=PC2))
  p <- p + geom_point(aes(color = Gender, shape=Reprogramming_Vector_Type, size=Cell_Line_Type))
  p <- p + theme_bw() + scale_size_manual(values = c(5, 2)) + scale_shape_manual(values = 1:7) + theme(legend.position="top") 
  p <- p + geom_text(aes(label= labels), size=4, hjust=0)
  print(p)
  
  return(list(p = p))
}

# PCA plots for methyl expression
PCA.METHYL.EXPRESSION = lapply(COVARIATES, plotPCAMethylExpression, PROC.EXP)
```
### Significant Covariates
Correlation between pca of unadjusted expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, include = TRUE, cache=FALSE, fig.width=25, fig.height=12, cache = FALSE}
plotMethylCovariatesCorrelation <- function(COVARIATES, PROC.EXP){
  writeLines('------------------------------------------------------------------------------')
  diffstate = unique(COVARIATES$Diffname_short)
  writeLines(paste('Differentiation state:',as.character(diffstate)))
  writeLines('------------------------------------------------------------------------------')
  
  # Drop unilevel covariates
  COVARIATES = droplevels(COVARIATES)
  COVARIATES = COVARIATES[,c(which(sapply(COVARIATES, function(x) length(levels(x))) > 1),
                             which(colnames(COVARIATES) %in% ContCovariates))]    
  
  # Find correlation between PC's of gene expression with covariates
  DESIGN = getDesignMatrix(COVARIATES, Intercept = F)
  colnames(DESIGN$design) = sapply(colnames(DESIGN$design),function(x){substr(x,1,50)})
  
  preAdjustedSigCovars = runPCAandPlotCorrelations(PROC.EXP[[diffstate]], 
                                                   data.matrix(COVARIATES),
                                                   'NULL design')
  
  # Find significant covariates
  adjustCovars = preAdjustedSigCovars$significantCovars
  
  # Plot preAdjustedSigCovars
  print(preAdjustedSigCovars[["PC_res"]][[2]]$plotData)  
  
  writeLines('------------------------------------------------------------------------------')
  writeLines(paste('Significant covariates to adjust at FDR 0.1 are',paste(gsub('_','_',adjustCovars), collapse= ',')))
  writeLines('------------------------------------------------------------------------------')
  writeLines('')
  writeLines('')
  
  return(list(DESIGN = DESIGN,
              adjustCovars = adjustCovars,
              preAdjustedSigCovars = preAdjustedSigCovars))  
}

PREADJUSTED.COVARIATES = lapply(COVARIATES, plotMethylCovariatesCorrelation, PROC.EXP)
```
### Normalisation (iterative)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix. All effects are considered as fixed effects, except Donor_ID which is a random effect
```{r iterative.adjusted.normalisation, cache = FALSE, include=TRUE, fig.width=25, fig.height=12}
iterativeMethylNormalisation <- function(COVARIATES, 
                                         PROC.EXP,
                                         PREADJUSTED.COVARIATES){
  writeLines('------------------------------------------------------------------------------')
  diffstate = unique(COVARIATES$Diffname_short)  
  writeLines(paste('Differentiation state:',as.character(diffstate)))
  writeLines('------------------------------------------------------------------------------')
  
  # Drop unilevel covariates
  COVARIATES = droplevels(COVARIATES)
  COVARIATES = COVARIATES[,c(which(sapply(COVARIATES, function(x) length(levels(x))) > 1),
                             which(colnames(COVARIATES) %in% ContCovariates))]    
  
  # Assign return values
  FIT = c()
  CORRELATION = c()
  RESIDUAL.METHYL.EXPRESSION = c()  
  postAdjustCovars = c()
  
  # Assign residual covariates
  residualSigCovars = PREADJUSTED.COVARIATES[[diffstate]]$preAdjustedSigCovars
  
  significantCovars = residualSigCovars$significantCovars
  covariatesEffects = residualSigCovars$Effects.significantCovars
  covariatesEffects = covariatesEffects[setdiff(significantCovars,c(postAdjustCovars,'Donor_ID'))]
  
  postAdjustCovars = names(which.max(covariatesEffects))
  
  loopCount = 0 
  while(length(covariatesEffects) != 0 && loopCount <= 15){
    writeLines(paste('Using following covariates in the model:',
                     paste(postAdjustCovars,collapse=', '),
                     'as fixed effects'
    ))
    
    # Post adjusted design matrix
    DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
    DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
    
    # Estimate correlation between donor effects
    CORRELATION = duplicateCorrelation(PROC.EXP[[diffstate]],
                                       design = DM1$design,
                                       block = COVARIATES$Donor_ID)
    
    # CORRELATION$consensus.cor = 0.2
    if(abs(CORRELATION$cor) == 1 || abs(CORRELATION$cor) == 0){
      writeLines('Random effects correlation cannot be estimated')
      break
    }
    
    # Fit linear model with chosen design
    FIT = lmFit(PROC.EXP[[diffstate]],
                design = DM1$design,
                block = COVARIATES$Donor_ID,
                correlation = CORRELATION$consensus.cor)
    
    # Residuals after normalisation
    RESIDUAL.METHYL.EXPRESSION = residuals.MArrayLM(FIT,PROC.EXP[[diffstate]])
    
    # Residual covariates to choose from
    residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, 'Donor_ID'))
    
    # Find PC of residual gene expression and significant covariates that are highly correlated with PCs 
    DESIGN = as.data.frame(PREADJUSTED.COVARIATES[[diffstate]]$DESIGN$design)
    residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.METHYL.EXPRESSION, 
                                                  data.matrix(COVARIATES),
                                                  'residuals')
    
    # Add postadjusted covariates (if any)
    significantCovars = residualSigCovars$significantCovars
    covariatesEffects = residualSigCovars$Effects.significantCovars
    covariatesEffects = covariatesEffects[setdiff(significantCovars,c(postAdjustCovars,'Donor_ID'))]
    
    postAdjustCovars = c(postAdjustCovars,names(which.max(covariatesEffects)))
    write.table(diffstate, file = './temp.txt', append=T, col.names=F, row.names=F,quote=F)
    write.table(paste(postAdjustCovars, collapse=','), file = './temp.txt', append=T, col.names=F, row.names=F,quote=F)
    loopCount = loopCount + 1
  }
  writeLines('--------------------------------------------------------------------------------------------')
  writeLines(paste('Using following covariates in the final model', paste(paste(postAdjustCovars,collapse=','),'as fixed effects and Donor_ID as random effects')))
  writeLines('--------------------------------------------------------------------------------------------')
  writeLines('')
  
  ### Sanity Check
  print(residualSigCovars[["PC_res"]][[2]]$plotData)
  
  writeLines('---------------------------------------------------------------------------------------------')
  
  return(list(FIT = FIT,
              RESIDUAL.METHYL.EXPRESSION = RESIDUAL.METHYL.EXPRESSION,
              residualSigCovars = residualSigCovars,
              postAdjustCovars = postAdjustCovars))
}
POSTADJUSTED.COVARIATES = lapply(COVARIATES, iterativeMethylNormalisation, PROC.EXP, PREADJUSTED.COVARIATES)
```

### Synapse Store
Store processed beta values, covariates, and design matrices
```{r synapsestore, echo=FALSE, include=FALSE, eval=TRUE, cache=FALSE}
ActivityName <- 'Covariate Analysis of minfi aligned methylation data at each diff state with mixed effects'

thisFileName <- 'CovariateAnalysis_methylation_mixedEffects_EachDiffState_withECTOScores.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/pcbc_c4_analysis", 
                    ref="branch", 
                    refName='methyl')
thisFile <- getPermlink(repository = thisRepo, 
                        repositoryPath=paste0('code/Rmd/', thisFileName))

# Populate wiki with results
CODE <- Folder(name = 'Covariate Analysis Methylation Mixed Effects EachDiffState withECTOScores', 
               parentId = parentId)
CODE <- synStore(CODE)

storeFilesInSynapse <- function(COVARIATES, PROC.EXP, POSTADJUSTED.COVARIATES, CODE){
  diffstate = unique(COVARIATES$Diffname_short)  
  
  # Drop unilevel covariates
  COVARIATES = droplevels(COVARIATES)
  COVARIATES = COVARIATES[,sapply(COVARIATES, function(x) length(levels(x))) > 1]    
  
  # Store expression
  EXP <- rownameToFirstColumn(PROC.EXP[[diffstate]],'ProbeIDs')
  write.table(EXP, 
              file = paste0('./Filtered_Beta_',diffstate,'.tsv'), 
              sep='\t', 
              row.names=F, 
              col.names=T, 
              quote=F)
  EXP_OBJ <- File(paste0('./Filtered_Beta_',diffstate,'.tsv'), 
                  name = paste('Filtered Beta Values',diffstate), 
                  parentId = CODE$properties$id)
  EXP_OBJ <- synStore(EXP_OBJ, 
                      used = ALL_USED_IDs, 
                      activityName = ActivityName, 
                      executed = thisFile) 
  
  # Covariates Matrix
  COVARIATES <- rownameToFirstColumn(COVARIATES,'UID')
  write.table(COVARIATES, 
              file = paste0('./Adjusted_Methylation_Covariates_',diffstate,'.tsv'), 
              sep='\t', 
              row.names=F, 
              col.names=T,
              quote=F)
  COVARIATES_OBJ <- File(paste0('Adjusted_Methylation_Covariates_',diffstate,'.tsv'),
                         name = paste('Adjusted Covariates',diffstate),
                         parentId = CODE$properties$id)
  COVARIATES_OBJ <- synStore(COVARIATES_OBJ, 
                             used = ALL_USED_IDs, 
                             activityName = ActivityName, 
                             executed = thisFile)
  
  # Design Matrix
  DESIGN <- rownameToFirstColumn(POSTADJUSTED.COVARIATES[[diffstate]]$FIT$design,'UID')
  write.table(DESIGN,
              file = paste0('./Adjusted_Methylation_Design_',diffstate,'.tsv'),
              sep='\t',
              row.names=F,
              col.names=T,
              quote=F)
  DESIGN_OBJ <- File(paste0('Adjusted_Methylation_Design_',diffstate,'.tsv'),
                     name = paste('Adjusted Design',diffstate),
                     parentId = CODE$properties$id)
  DESIGN_OBJ <- synStore(DESIGN_OBJ,
                         used = ALL_USED_IDs, 
                         activityName = ActivityName, 
                         executed = thisFile)
}
STORED.FILES = lapply(COVARIATES, storeFilesInSynapse, PROC.EXP, POSTADJUSTED.COVARIATES, CODE)
```
### Executed: Source code
`r paste0('[Source R Markdown](',print(thisFile),')')`