---
title: "Covariate Analysis for eXpress aligned data with mixed effect modeling"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file is

# Clear R console screen output
cat("\014")  

# Load required libraries
library('synapseClient')
library('RColorBrewer')
library('ggplot2')
library('gplots')
library('limma')
library('edgeR')
library('ctv')
library('psych')
library('reshape2')
library('vcd')
library('erer')
library('fpc')
library('data.table')
library('dplyr')
library('knitr')
library('stringr')

## Requires ggdendro
devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")

# source('/mnt/Github/knit2synapse-1/R/knitFile2Synapse.R')
# knitFile2Synapse(file = "./CovariateAnalysis_Express_mixedEffects.Rmd", name = 'Covariate Analysis Express Mixed Effects', owner = 'syn3943944', overwrite=F)

synapseLogin()

# source utility files from ../R/lib folder
file.sources = list.files('../R/lib',pattern="*.R")
file.sources = sapply(file.sources,function(x){return(paste('../R/lib',x,sep='/'))})
tmp = sapply(file.sources,source,.GlobalEnv)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE)
```

### Download data

```{r setup, include=FALSE}
# Input Parameters
COUNT_ID = 'syn3446250'
METADATA_ID = 'syn3156503'

SYNAPSE_STORE = T
parentId = 'syn3276099'

# Specify factor and continuous covarites pool (adjusted covariates will only be the subset of these covariates)
FactorCovariates = c('Diffname_short','run','lane','index','Cell_Type_of_Origin', 'Cell_Line_of_Origin', 'Tissue_of_Origin', 'Reprogramming_Vector_Type','Reprogramming_Gene_Combination', 'Donor_Life_Stage','Originating_Lab','Gender','Donor_ID','Disease','Race','Ethnicity','Cell_Line_Type','PassageAtHarvest')
ContCovariates = NULL
```
Synapse id of count matrix used for the analysis is `r COUNT_ID` and the synapse id of meta data table used for the analysis is `r METADATA_ID`. 

Factor covariates considered for analysis are `r FactorCovariates`, and continuous covariates considered for the analysis are `r ContCovariates`.

Obtain count matrix and metadata from synapse.
```{r getdata, cache=TRUE, include=FALSE}
# Get count matrix
COUNT_OBJ = synGet(COUNT_ID)
ALL_USED_IDs = COUNT_OBJ$properties$id
COUNT = fread(getFileLocation(COUNT_OBJ), data.table=FALSE)
row.names(COUNT) = COUNT[,1]

# Get metadata
METADATA_OBJ = synTableQuery(paste('SELECT * FROM',METADATA_ID,sep=' '))
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_OBJ@schema
METADATA = METADATA_OBJ@values
```

Preprocess counts matrix and metadata.
We separate MESO samples into early (5 day) and late (15 and 30 day).
```{r preprocessing, include=FALSE}
# Preprocess metadata
METADATA[METADATA == 'N/A'] = 'Not Applicable'

# Preprocess passage at harvest information
METADATA$PassageAtHarvest[is.na(METADATA$PassageAtHarvest)] = 'Not Available' 

# Assign new values for MESO-5, 15 and 30 samples
METADATA$Diffname_short[METADATA$Diffname_short == 'MESO-5'] = 'MESO_EARLY'
METADATA$Diffname_short[METADATA$Diffname_short == 'MESO-15' | 
                          METADATA$Diffname_short == 'MESO-30'] = 'MESO_LATE'
rownames(METADATA) = METADATA$UID

```

### Preprocess data
* Remove somatic samples and samples with type NA.
* Remove samples that failed QC and samples classified as exclude.
* Remove samples with abnormal karyotypes.
```{r filtering, echo=TRUE}
#### Pre processing mRNA expression counts and metadata ####
metadata_filtered <- 
  METADATA %>%
  filter(Diffname_short != "") %>%
  filter(UID %in% colnames(COUNT)) %>%
  filter(Cell_Type == "PSC") %>%
  filter(!is.na(Cell_Type)) %>%
  filter(pass_qc) %>%
  filter(!exclude) %>%
  filter(C4_Karyotype_Result != "abnormal" | !is.na(C4_Karyotype_Result))

REMOVED_UID <- setdiff(METADATA$UID, metadata_filtered$UID)
METADATA <- METADATA[metadata_filtered$UID,]
COUNT <- COUNT[, METADATA$UID]
```

The following `r length(REMOVED_UID)` samples were removed:

`r REMOVED_UID` 

### CPM Normalisation
Determine design matrix for normalisation and differential expression analysis. 

Remove genes that have less than 1 cpm counts in at least 50% of each of the individual differentiation stages.
```{r cpmnormalisation}
tmp <- tapply(colnames(COUNT),
              factor(METADATA$Diffname_short),
              function(cols,COUNT){PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[,cols])},
              COUNT)

ugenes <- c()
for (i in unique(METADATA$Diffname_short)) {
  ugenes <- unique(c(ugenes,tmp[[i]]$filteredExprMatrix$genes[,1]))
}

COUNT <- COUNT[ugenes,,drop=F]
PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT,MIN_GENE_CPM=0,
                                                 MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
```

### Covariate clustering
Determine relationship between covariates. 

```{r covariates.clustering}
# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates)]

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
```

Covariate correlation

```{r covariates.correlation, fig.width=10, fig.height=10}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 1)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.3, row.width=0.15)
```

### Normalisation (with NULL)
Initial normalisation usign voom (with NULL design)

```{r initial.voom.normalisation}
# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=NULL, plot=T)

# Initial design matrix with covariates under consideration
DM = getDesignMatrix(COVARIATES, Intercept = F)
```

Clustering of initial normalised data (with NULL design)

```{r decompse.normalise.data, cache=TRUE, fig.height=10, fig.width=10}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E)

# Determine number of clusters automatically using pamk
pam.cluster <- pamk(t(VOOM.GENE_EXPRESSION$E), krange=2:10)

# Plot first 2 PCs
plotdata <- data.frame(UID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2], 
                       cluster=pam.cluster$pamobject$clustering)

plotdata <- merge(plotdata, METADATA, by="UID")
plotdata <- mutate(plotdata, labels = '');#ifelse(Diffname_short == '',as.character(UID),'')) 

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=factor(cluster), shape=Diffname_short, size=Gender))
p <- p + theme_bw() + scale_size_manual(values = c(5, 2)) + theme(legend.position="top") 
p <- p + geom_text(aes(label= labels), size=4, hjust=0)
p
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates}
# Find correlation between PC's of gene expression with covariates
DESIGN = getDesignMat2Fact(DM$design,FactorCovariates)
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, DESIGN,'NULL design(voom-normalized)', isKeyPlot=TRUE)

# Find significant covariates
adjustCovars = designMatVarsToCovars(getCovariatesMap(DM),preAdjustedSigCovars$significantCovars)
```

Significant covariates to adjust at FDR 0.1 are `r adjustCovars`

```{r preAdjustedSigCovars.NULL, fig.width=15, fig.height=8}
preAdjustedSigCovars[["PC_res"]][[1]]$plotData
```

```{r preAdjustedSigCovars.NULL.ALL, fig.width=15, fig.height=8}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

### Normalisation (iterative)
Since many covariates are confounded, re-normalising COVARIATES with an iterative design matrix. Here Donor_ID is chosen as random effect

```{r iterative.adjusted.voom.normalisation}
postAdjustCovars = list()
postAdjustCovars$random = 'Donor_ID'
 
covariatesEffects = designMat2CovEffects(getCovariatesMap(DM),preAdjustedSigCovars$Effects.significantCovars)
covariatesEffects = covariatesEffects[setdiff(names(covariatesEffects),postAdjustCovars$random)]

postAdjustCovars$fixed = names(which.max(covariatesEffects))
residualSigCovars = preAdjustedSigCovars

loopCount = 0 
while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
  print(paste('Using following covariates in the model',
              paste(paste(postAdjustCovars$fixed,collapse=','),'as fixed effects and', paste(postAdjustCovars$random,collapse=','),'as random effects'),
              sep=':'))
  
  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars$fixed,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
    
  # Estimate correlation of random effects
  VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F)
  correlation <- duplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION$E, design=DM1$design, block=COVARIATES$Donor_ID)
  
  # Re-calculate voom weights with correlation of random effects
  VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F,
                                       block=COVARIATES$Donor_ID, correlation = correlation$cor)
  
  tmp = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION$E, design=DM1$design, weights = VOOM.ADJUSTED.GENE_EXPRESSION$weights,
                                        block=COVARIATES$Donor_ID, correlation = correlation$cor)
  
  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(tmp,VOOM.ADJUSTED.GENE_EXPRESSION$E)
  
  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates), postAdjustCovars$fixed)
  
  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                dplyr::select(DESIGN,-starts_with('Donor_ID')),
                                                'all adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)
  
  # Add postadjusted covariates (if any)
  covariatesEffects = designMat2CovEffects(getCovariatesMap(DM),residualSigCovars$Effects.significantCovars)
  covariatesEffects = covariatesEffects[setdiff(names(covariatesEffects),postAdjustCovars$random)]
  
  postAdjustCovars$fixed = c(postAdjustCovars$fixed,names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}
tmp <- paste('Using following covariates in the final model', paste(paste(postAdjustCovars$fixed,collapse=','),'as fixed effects and', paste(postAdjustCovars$random,collapse=','),'as random effects'))
```

`r tmp`

Degenerate variables are: 
```{r adjusted.voom.normalisation}
# Post adjusted design matrix
DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars$fixed,drop=F], Intercept = F)
tmp = linColumnFinder(DM1$design)
print(tmp$relations)
DM1$design = DM1$design[,tmp$indepCols] 

# Recalculate weights with new design
VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F)
correlation <- duplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION$E, design=DM1$design, block=COVARIATES$Donor_ID)

# Re-calculate voom weights with correlation of random effects
VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=T,
                                     block=COVARIATES$Donor_ID, correlation = correlation$cor)

# Linear model with new weights and design
VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION$E, design=DM1$design, weights = VOOM.ADJUSTED.GENE_EXPRESSION$weights,
                          block=COVARIATES$Donor_ID, correlation = correlation$cor)
  
```

### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates

```{r calculate.residuals, include=FALSE}
# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              dplyr::select(DESIGN,-starts_with('Donor_ID')),
                                              'residual matrix of all adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```

```{r residualSigCovars, fig.width=15, fig.height=8}
residualSigCovars[["PC_res"]][[1]]$plotData
```

```{r residualSigCovars.ALL, fig.width=15, fig.height=8}
residualSigCovars[["PC_res"]][[2]]$plotData
```

### Normalisation (custom/manual design)
Reperform normalisation with custom design matrix (manually with Diffname short, run, Gender, Donor Life Stage, and Reprogramming Vector Type as fixed effects and Donor ID as random effects)

```{r reperform.normalisation, cache=TRUE}
# Post adjusted design matrix
DM1 = getDesignMatrix(COVARIATES[,c("Diffname_short","run","Gender","Donor_Life_Stage","Reprogramming_Vector_Type"),drop=F],Intercept = F)
DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
    
# Estimate correlation of random effects
VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F)
correlation <- duplicateCorrelation(VOOM.ADJUSTED.GENE_EXPRESSION$E, design=DM1$design, block=COVARIATES$Donor_ID)
  
# Re-calculate voom weights with correlation of random effects
VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=DM1$design, plot=F,
                                     block=COVARIATES$Donor_ID, correlation = correlation$cor)
  
# Re-calculate linear fit using new weights and new design
VOOM.ADJUSTED.FIT = lmFit(VOOM.ADJUSTED.GENE_EXPRESSION$E, design=DM1$design, weights = VOOM.ADJUSTED.GENE_EXPRESSION$weights,
                          block=COVARIATES$Donor_ID, correlation = correlation$cor)

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION,
                                              dplyr::select(DESIGN,-starts_with('Donor_ID')),
                                              'all adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```

```{r residualSigCovars.manual, fig.width=15, fig.height=8}
residualSigCovars[["PC_res"]][[1]]$plotData
```

```{r residualSigCovars.manual.ALL, fig.width=15, fig.height=8}
residualSigCovars[["PC_res"]][[2]]$plotData
```

Readjust for covariates, normalise and perform outlier analysis (as a sanity check)

```{r new.design, cache=TRUE, fig.height=10, fig.width=10}
# Find principal components of expression to plot
PC <- prcomp(VOOM.ADJUSTED.GENE_EXPRESSION$E)

# Determine number of clusters automatically using pamk
pam.cluster <- pamk(t(VOOM.ADJUSTED.GENE_EXPRESSION$E), krange=2:10)

# Plot first 2 PCs
plotdata <- data.frame(UID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2], 
                       cluster=pam.cluster$pamobject$clustering)

plotdata <- mutate(plotdata, labels = "") 

plotdata <- merge(plotdata, METADATA, by="UID")

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=factor(cluster), shape=Diffname_short, size=Gender))
p <- p + theme_bw() + scale_size_manual(values = c(5, 2)) + theme(legend.position="top") 
p <- p + geom_text(aes(label= labels), size=4, hjust=0)
p
```

Store covariates, design matrix, and voom adjusted gene expression

```{r synapsestore, echo=FALSE, include=FALSE, eval=TRUE}
if (SYNAPSE_STORE){  
  ActivityName <- 'Covariate Analysis of eXpress aligned data with mixed effects modeling'
  
  # Code
  CODE <- File('./CovariateAnalysis_Express_mixedEffects.Rmd',name = 'Covariate Analysis Express Mixed Effects',parentId = parentId)
  CODE <- synStore(CODE, used = ALL_USED_IDs, activityName = ActivityName, executed = 'https://github.com/th1vairam/pcbc_c4_analysis/blob/afc96cd1a43ed9bb4b7fafbbcd20322991630f78/code/Rmd/CovariateAnalysis_Express_mixedEffects.Rmd')
  
  # Expression Data
  EXP <- rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E,'GeneName')  
  write.table(EXP,file = './AdjustedExpression_Express_MixedEffects.tsv',sep='\t',row.names=F,col.names=T,quote=F)
  EXPRESSION <- File('./AdjustedExpression_Express_MixedEffects.tsv',name = 'Adjusted Expression Express Mixed Effects',parentId = parentId)
  EXPRESSION <- synStore(EXPRESSION, used = ALL_USED_IDs, activityName = ActivityName, executed = CODE$properties$id)  
  
  # Adjustment Weights 
  WEIGHTS <- cbind(EXP$GeneName,VOOM.ADJUSTED.GENE_EXPRESSION$weights)
  colnames(WEIGHTS) <- colnames(EXP)  
  write.table(WEIGHTS,file = './AdjustedWeights_Express_MixedEffects.tsv',sep='\t',row.names=F,col.names=T,quote=F)
  WEIGHTS <- File('AdjustedWeights_Express_MixedEffects.tsv',name = 'Adjusted Weights Express Mixed Effects',parentId = parentId)
  WEIGHTS <- synStore(WEIGHTS, used = ALL_USED_IDs, activityName = ActivityName, executed = CODE$properties$id)
  
  # Covariates Matrix
  COVARIATES <- rownameToFirstColumn(COVARIATES,'UID')
  write.table(COVARIATES,file = './Covariates_Express_MixedEffects.tsv',sep='\t',row.names=F,col.names=T,quote=F)
  COVARIATES <- File('Covariates_Express_MixedEffects.tsv',name = 'Covariates Express Mixed Effects',parentId = parentId)
  COVARIATES <- synStore(COVARIATES, used = ALL_USED_IDs, activityName = ActivityName, executed = CODE$properties$id)
  
  # Design Matrix
  DESIGN <- rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$design,'UID')
  write.table(DESIGN,file = './AdjustedDesign_Express_MixedEffects.tsv',sep='\t',row.names=F,col.names=T,quote=F)
  DESIGN <- File('AdjustedDesign_Express_MixedEffects.tsv',name = 'Adjusted Design Express Mixed Effects',parentId = parentId)
  DESIGN <- synStore(DESIGN, used = ALL_USED_IDs, activityName = ActivityName, executed = CODE$properties$id)
}
```
|  *Results*      |  *SynapseID*                  |
|  -------        |   ---------                   |
|  Expression     |  `r EXPRESSION$properties$id` |
|  sample weights |  `r WEIGHTS$properties$id`    |
|  Covariates     |  `r COVARIATES$properties$id` |
|  Design Matrix  |  `r DESIGN$properties$id`     |