---
title: "Differential splicing analysis for Alt-Analyze data with mixed effects modeling"
author: "Thanneer Perumal"
date: "`r date()`"
---
```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file is

# Clear R console screen output
cat("\014")  

# Load required libraries
library('RColorBrewer')
library('ggplot2')

library('psych')
library('reshape2')

library('limma')
library('vcd')
library('data.table')
library('plyr')
library('dplyr')

library('knitr')
library('stringr')
library('synapseClient')
library('rGithubClient') ## Needs the dev branch

## Requires ggdendro
devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")

synapseLogin()
# 
# knit2synapse::knitToFolderEntity(file = "./DiffExpAnal_splicing_mixedEffects_All.Rmd", 
#                                  entityName = 'Differential Splicing Analysis Mixed Effects All', 
#                                  parentId = 'syn4991628', 
#                                  overwrite=F, knitmd = T)

# source utility files from ../R/lib folder
file.sources = list.files('../R/lib',pattern="*.R", full.names = TRUE)
tmp = sapply(file.sources,source,.GlobalEnv)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE)
```

```{r fxns, include=FALSE}
# Function to rotate variables
rotateVars <- function(x, first) {
  others <- setdiff(x, first)
  c(first, others)
}

# Function to make design matrix
makeDesignMatrix <- function(covariates, comparison, covariateColOrder=NA) {
  # Make a design matrix by reordering the column covariates
  # and column-binding each separate matrix for each column together
  # with complete listing of dummy variables (no reference levels)
  # and also remove any linearly dependent columns
  
  vars <- ifelse(is.na(covariateColOrder), 
                 colnames(covariates),
                 covariateColOrder)
  
  newVars <- rotateVars(vars, comparison)
  
  DESIGN = getDesignMatrix(covariates[, newVars], Intercept = F)
  DESIGN = DESIGN$design[, linColumnFinder(DESIGN$design)$indepCols]
  
  return(DESIGN)
}

# Perform differential splicing
doDiffExpr <- function(countMat, designMat, weights, block, comparison) {
  # Helper function to do all pairwise differential splicing on a single covariate
  # from the whole covariate matrix
  
  # Calculate correlation between donors
  correlation <- duplicateCorrelation(countMat, 
                                      design = designMat,
                                      weights = weights,
                                      block=block)
  
  # Fit linear model using mixed effects design
  fit = lmFit(countMat, 
              design = designMat, 
              weights = weights, 
              block=block, 
              correlation = correlation$cor)
  
  contrastNames <- colnames(designMat)[grep(comparison, colnames(designMat))]
  contrastNames <- combn(contrastNames, 2)
  contrastNames <- apply(contrastNames, 2, paste, collapse='-')
  
  cont <- makeContrasts(contrasts=contrastNames, 
                        levels=colnames(fit$coefficients))
  
  colnames(cont) <- sapply(colnames(cont),
                           function(x){x <- gsub(comparison, '', x);
                                       x <- gsub('-','_vs_',x);
                                       x <- paste('All',x,sep='__')})
  
  # Refit contrasts
  fitContrast <- contrasts.fit(fit, cont)
  
  # Estimate moderated t-statistics
  fitContrast <- eBayes(fitContrast)
  
  # Obtain all the differential expession combinations
  foo <- list()
  foo$logFC <- data.frame(row.names = rownames(countMat))
  foo$adj.P.Val <- data.frame(row.names = rownames(countMat))
  foo$SIG.SETS <- data.frame()
  
  for (i in colnames(cont)){
    tmp <- topTable(fitContrast, coef=i, number=dim(countMat)[1])    
    foo$logFC[,i] <- tmp[rownames(foo$logFC),'logFC']
    foo$adj.P.Val[,i] <- tmp[rownames(foo$adj.P.Val),'adj.P.Val'] 
    
    foo$SIG.SETS <- rbind(foo$SIG.SETS,
                          getUpDownGenes(foo$adj.P.Val[,i], foo$logFC[,i], 
                                         rownames(foo$logFC), i, FC_CUTOFF = 0))
  }
  
  # Get number of significantly differentialy spliced junctions with adj.P.Val <= 0.05 and logFC >= 0
  foo$SIG.EXP.POS <- foo$adj.P.Val<=0.05 & foo$logFC >= 0
  foo$NUM.SIG.EXP.POS <- colSums(foo$SIG.EXP.POS)
  
  # Get number of significantly differentialy spliced junctions with adj.P.Val <= 0.05 and logFC <= 0
  foo$SIG.EXP.NEG <- foo$adj.P.Val<=0.05 & foo$logFC <= 0
  foo$NUM.SIG.EXP.NEG <- colSums(foo$SIG.EXP.NEG)
  
  return(foo)
}
```

### Download data
```{r setup, include=FALSE}
# Input Parameters
PROCESSED_PSI_ID = 'syn5048714'
DESIGN_ID = 'syn5048721'
COVARIATES_ID = 'syn5048719'

SYNAPSE_STORE = T
parentId = 'syn4991628'

# Specify factor and continuous covariates pool (adjusted covariates will only be the subset of these covariates)
FactorCovariates = c('Diffname_short', 'run', 'lane', 'Cell_Line_Type', 'Cell_Line_of_Origin', 'Tissue_of_Origin', 'Reprogramming_Gene_Combination', 'Culture_Conditions', 'Donor_Life_Stage', 'Race', 'Ethnicity' , 'Gender', 'Disease', 'Originating_Lab', 'Donor_ID', 'Cell_Type_of_Origin_Level2', 'Reprogramming_Vector_Type')

ContCovariates = c('PassageAtThaw', 'PassageAtHarvest')

# The variables determined in the covariate analysis step
postAdjustedCovars <- c("Diffname_short", "Cell_Line_Type", "Gender", 
                        "Cell_Type_of_Origin_Level2",
                        "Donor_Life_Stage", "lane",
                        "Reprogramming_Gene_Combination",
                        "Culture_Conditions", "run")
```
Obtain processed PSI, design, and covariates matrix from Synapse.
```{r getdata, cache=TRUE, include=FALSE}
# Get PSI matrix
PROC_PSI_OBJ = synGet(PROCESSED_PSI_ID)
ALL_USED_IDs = PROC_PSI_OBJ$properties$id
PROC_PSI = fread(getFileLocation(PROC_PSI_OBJ), data.table=FALSE)

# Seperate PROC_PSI and PROC_PSI annotation
PSI.ANNOT = dplyr::select(PROC_PSI,  one_of(c("Symbol", "Description", "Minor-Isoform", "Major Isoform", "AltExons", "PME",
                                        "dPSI", "rho", "Max Inclusion PSI", "Coordinates", "feature")))
PROC_PSI = PROC_PSI[, setdiff(colnames(PROC_PSI), colnames(PSI.ANNOT))]
rownames(PROC_PSI) = PSI.ANNOT$"Minor-Isoform"

# Get sample covariates
COVARIATES_OBJ = synGet(COVARIATES_ID)
ALL_USED_IDs = c(ALL_USED_IDs, COVARIATES_OBJ$properties$id)
COVARIATES = fread(getFileLocation(COVARIATES_OBJ), data.table=FALSE)
row.names(COVARIATES) = COVARIATES[,1]
COVARIATES = COVARIATES[,-(1)]

# Convert factor covariates to factors
FactorCovariates = intersect(colnames(COVARIATES), FactorCovariates)
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
ContCovariates = intersect(colnames(COVARIATES), ContCovariates)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
```
Synapse IDs of used data:
  | **Name**            | **SynapseID** |
  | Processed PSI       | `r paste(PROC_PSI_OBJ$properties$id,PROC_PSI_OBJ$properties$versionNumber,sep='.')` |
  | Covariates (subset) | `r paste(COVARIATES_OBJ$properties$id,COVARIATES_OBJ$properties$versionNumber,sep='.')` |
  
### Differential splicing analysis
For each covariate, we fit a linear model using `limma` with only the variables obtained from the covariate analysis:

  * Fixed Effects<br/>
    `r paste(gsub('_','\\\\_', postAdjustedCovars), collapse=", ")`
  * Random Effects<br/>
    Donor\_ID

The design matrix is formed by including all variables for each factor encoded as dummy variables, reordered so that the variable of interest for comparison is first. With this formulation and the available data, many design matrices are degenerate. Hence, we also remove any columns that are linear combinations of other columns. Because of this, models between separate covariates (separate columns in the covariate matrix) cannot be compared, as each comparison is performed with a separately defined model.

#### Differentiation stages
```{r diffstate, cache=TRUE, include=FALSE}
comparison <- "Diffname_short"

DESIGN <- makeDesignMatrix(covariates=COVARIATES, comparison = comparison, 
                           covariateColOrder = postAdjustedCovars)

DIFFNAME <- doDiffExpr(countMat=PROC_PSI, designMat=DESIGN,
                       weights=NULL, block=COVARIATES$Donor_ID,
                       comparison=comparison)
```
Get differentially spliced junctions using limma package with following coefficients in the linear model:
  
  `r paste(gsub('_','\\\\_',colnames(DESIGN)),collapse=', ')`

Number of differentially spliced junctions at FDR <= 0.05 and logFC >= 0 or logFC <= 0:
```{r significant.sets1,cache=TRUE}
tmp <- cbind(as.data.frame(DIFFNAME$NUM.SIG.EXP.POS),
             as.data.frame(DIFFNAME$NUM.SIG.EXP.NEG))

colnames(tmp) <- c('NUMBER OF UP REG. SPLICE JN.','NUMBER OF DOWN REG. SPLICE JN.')
kable(tmp)
```
#### Cell line type
```{r celllinetype, cache=TRUE, include=FALSE}
comparison <- "Cell_Line_Type"

DESIGN <- makeDesignMatrix(covariates=COVARIATES, comparison = comparison, 
                           covariateColOrder = postAdjustedCovars)

CELL.LINE.TYPE <- doDiffExpr(countMat=PROC_PSI, designMat=DESIGN,
                             weights=NULL, block=COVARIATES$Donor_ID,
                             comparison=comparison)
```
Get differentially spliced junctions using limma package with following coefficients in the linear model:
  
  `r paste(gsub('_','\\\\_',colnames(DESIGN)),collapse=', ')`

Number of differentially spliced junctions at FDR <= 0.05 and logFC >= 0 or logFC <= 0:
```{r significant.celllinetype,cache=TRUE}
tmp <- cbind(as.data.frame(CELL.LINE.TYPE$NUM.SIG.EXP.POS),
             as.data.frame(CELL.LINE.TYPE$NUM.SIG.EXP.NEG))

colnames(tmp) <- c('NUMBER OF UP REG. SPLICE JN.','NUMBER OF DOWN REG. SPLICE JN.')
kable(tmp)
```
#### Gender
```{r gender, cache=TRUE}
comparison <- "Gender"

DESIGN <- makeDesignMatrix(covariates=COVARIATES, comparison = comparison, 
                           covariateColOrder = postAdjustedCovars)

GENDER <- doDiffExpr(countMat=PROC_PSI, designMat=DESIGN,
                     weights=NULL, block=COVARIATES$Donor_ID,
                     comparison=comparison)
```
Get differentially spliced junctions using limma package with following co-efficients in the linear model:
  
  `r paste(gsub('_','\\\\_',colnames(DESIGN)), collapse = ', ')`

Number of differentially spliced junctions at FDR <= 0.05 and absolute logFC >= 0 or logFC <= 0:
```{r significant.sets2,cache=TRUE}
tmp <- cbind(as.data.frame(GENDER$NUM.SIG.EXP.POS),
             as.data.frame(GENDER$NUM.SIG.EXP.NEG))

colnames(tmp) <- c('NUMBER OF UP REG. SPLICE JN.', 'NUMBER OF DOWN REG. SPLICE JN.')
kable(tmp)
```
#### Cell Type of Origin Level2
```{r cell.type.of.origin, cache=TRUE}
comparison <- "Cell_Type_of_Origin_Level2"

DESIGN <- makeDesignMatrix(covariates=COVARIATES, comparison = comparison, 
                           covariateColOrder = postAdjustedCovars)

CELL.TYPE.OF.ORIGIN <- doDiffExpr(countMat=PROC_PSI, designMat=DESIGN,
                                  weights=NULL, block=COVARIATES$Donor_ID,
                                  comparison=comparison)
```
Get differentially spliced junctions using limma package with following co-efficients in the linear model:
  
  `r paste(gsub('_','\\\\_',colnames(DESIGN)), collapes= ', ')`

Number of differentially spliced junctions at FDR <= 0.05 and logFC >= 0 or logFC <= 0:
```{r significant.sets3,cache=TRUE}
tmp <- cbind(as.data.frame(CELL.TYPE.OF.ORIGIN$NUM.SIG.EXP.POS),
             as.data.frame(CELL.TYPE.OF.ORIGIN$NUM.SIG.EXP.NEG))

colnames(tmp) <- c('NUMBER OF UP. REG. JN.', 'NUMBER OF DOWN REG. SPLICE JN.')
kable(tmp)
```
#### Donor Life Stage
```{r donorlifestage, cache=TRUE, include=FALSE}
comparison <- "Donor_Life_Stage"

DESIGN <- makeDesignMatrix(covariates=COVARIATES, comparison = comparison, 
                           covariateColOrder = postAdjustedCovars)

DONOR.LIFE.STAGE <- doDiffExpr(countMat=PROC_PSI, designMat=DESIGN,
                               weights=NULL, block=COVARIATES$Donor_ID,
                               comparison=comparison)
```
Get differentially spliced junctions using limma package with following coefficients in the linear model:
  
  `r paste(gsub('_','\\\\_',colnames(DESIGN)),collapse=', ')`

Number of differentially spliced junctions at FDR <= 0.05 and logFC >= 0 or logFC <= 0:
```{r significant.donorlifestage,cache=TRUE}
tmp <- cbind(as.data.frame(DONOR.LIFE.STAGE$NUM.SIG.EXP.POS),
             as.data.frame(DONOR.LIFE.STAGE$NUM.SIG.EXP.NEG))

colnames(tmp) <- c('NUMBER OF UP REG. SPLICE JN.','NUMBER OF DOWN REG. SPLICE JN.')
kable(tmp)
```
#### Reprogramming Gene Combination
```{r rep.gene.combo, cache=TRUE}
comparison <- "Reprogramming_Gene_Combination"

DESIGN <- makeDesignMatrix(covariates=COVARIATES, comparison = comparison, 
                           covariateColOrder = postAdjustedCovars)

REP.GENE.COMBO <- doDiffExpr(countMat=PROC_PSI, designMat=DESIGN,
                             weights=NULL, block=COVARIATES$Donor_ID,
                             comparison=comparison)
```
Get differentially spliced junctions using limma package with following co-efficients in the linear model:
  
  `r paste(gsub('_','\\\\_',colnames(DESIGN)), collapse=', ')`

Number of differentially spliced junctions at FDR <= 0.05 and logFC >= 0 or logFC <= 0:
```{r significant.sets6,cache=TRUE}
tmp <- cbind(as.data.frame(REP.GENE.COMBO$NUM.SIG.EXP.POS),
             as.data.frame(REP.GENE.COMBO$NUM.SIG.EXP.NEG))

colnames(tmp) <- c('NUMBER OF UP REG. SPLICE JN.','NUMBER OF DOWN REG. SPLICE JN.')
kable(tmp)
```
#### Culture Conditions
```{r culture.conditions, cache=TRUE}
comparison <- "Culture_Conditions"

DESIGN <- makeDesignMatrix(covariates=COVARIATES, comparison = comparison, 
                           covariateColOrder = postAdjustedCovars)

CULT.COND <- doDiffExpr(countMat=PROC_PSI, designMat=DESIGN,
                        weights=NULL, block=COVARIATES$Donor_ID,
                        comparison=comparison)
```
Get differentially spliced junctions using limma package with following co-efficients  in the linear model:
  
  `r paste(gsub('_','\\\\_',colnames(DESIGN)), collapse=', ')`

Number of differentially spliced junctions at FDR <= 0.05 and logFC >= 0 or logFC <= 0:
```{r significant.sets5,cache=TRUE}
tmp <- cbind(as.data.frame(CULT.COND$NUM.SIG.EXP.POS),
             as.data.frame(CULT.COND$NUM.SIG.EXP.NEG))

colnames(tmp) <- c('NUMBER OF UP REG. SPLICE JN.', 'NUMBER OF DOWN REG. SPLICE JN.')
kable(tmp)
```

```{r print.imp.JN.}
allFC <- llply(list(DIFFNAME$logFC, 
                    GENDER$logFC, 
                    CELL.LINE.TYPE$logFC,
                    CELL.TYPE.OF.ORIGIN$logFC,
                    CULT.COND$logFC, 
                    REP.GENE.COMBO$logFC,
                    DONOR.LIFE.STAGE$logFC),
               rownameToFirstColumn, colname="GeneNames")

FC <- plyr::join_all(allFC, by = 'GeneNames', match = 'all')

allPVal <- llply(list(DIFFNAME$adj.P.Val, 
                      GENDER$adj.P.Val, 
                      CELL.LINE.TYPE$adj.P.Val,
                      CELL.TYPE.OF.ORIGIN$adj.P.Val,
                      CULT.COND$adj.P.Val, 
                      REP.GENE.COMBO$adj.P.Val,
                      DONOR.LIFE.STAGE$adj.P.Val),
                 rownameToFirstColumn, colname="GeneNames")

PVAL <- join_all(allPVal, by = 'GeneNames', match = 'all')

allSig <- llply(list(DIFFNAME$SIG.EXP.POS + DIFFNAME$SIG.EXP.NEG,
                     GENDER$SIG.EXP.POS + GENDER$SIG.EXP.NEG,
                     CELL.LINE.TYPE$SIG.EXP.POS + CELL.LINE.TYPE$SIG.EXP.NEG,
                     CELL.TYPE.OF.ORIGIN$SIG.EXP.POS + CELL.TYPE.OF.ORIGIN$SIG.EXP.NEG,
                     CULT.COND$SIG.EXP.POS + CULT.COND$SIG.EXP.NEG,
                     REP.GENE.COMBO$SIG.EXP.POS + REP.GENE.COMBO$SIG.EXP.NEG,
                     DONOR.LIFE.STAGE$SIG.EXP.POS + DONOR.LIFE.STAGE$SIG.EXP.NEG),
                rownameToFirstColumn, colname="GeneNames")

SIG <- join_all(allSig, by = 'GeneNames', match = 'all')

SIG.SETS <- rbindlist(list(DIFFNAME$SIG.SETS,
                           CELL.LINE.TYPE$SIG.SETS,
                           GENDER$SIG.SETS,
                           CELL.TYPE.OF.ORIGIN$SIG.SETS,                  
                           CULT.COND$SIG.SETS,
                           REP.GENE.COMBO$SIG.SETS,
                           DONOR.LIFE.STAGE$SIG.SETS))
```

### Store files in Synapse
Store logFC, adjusted p-values and differentially spliced junctions.
```{r synapse.store, include = FALSE, eval=TRUE}
activityName='Differential splicing analysis of alt-analyse PSI with mixed effects model'

thisFileName <- './DiffExpAnal_splicing_mixedEffects_All.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/pcbc_c4_analysis", 
                    ref="branch", 
                    refName="splicing")

thisFile <- getPermlink(repository = thisRepo,
                        repositoryPath=paste0('code/Rmd/', thisFileName))

# Create folder to store the results and populate analysis wiki
CODE <- Folder(name = 'Differential Splicing Analysis Mixed Effects All', parentId = parentId)
CODE <- synStore(CODE)

# Store logFC
write.table(FC, file='./DiffExpAnal_splicing_mixedEffects_All_logFC.tsv',
            sep='\t',row.names=F,quote=F)

FC_OBJ <- File('./DiffExpAnal_splicing_mixedEffects_All_logFC.tsv',
               name = 'logFC', parentId = CODE$properties$id)
FC_OBJ <- synStore(FC_OBJ, used = ALL_USED_IDs, activityName = activityName, executed=thisFile)

# Store adjusted pvalue
write.table(PVAL,file='./DiffExpAnal_splicing_mixedEffects_All_adjPval.tsv',
            sep='\t',row.names=F,quote=F)

PVAL_OBJ <- File('./DiffExpAnal_splicing_mixedEffects_All_adjPval.tsv',
                 name = 'Adjusted Pvalue', parentId = CODE$properties$id)
PVAL_OBJ <- synStore(PVAL_OBJ, used = ALL_USED_IDs, activityName = activityName, executed=thisFile)

# Store differentially spliced junctions for comparison
write.table(SIG.SETS, file='./DiffExpAnal_splicing_mixedEffects_All_diffGenes.tsv',
            sep='\t', row.names=F, quote=F)

SIG.SETS_OBJ <- File('./DiffExpAnal_splicing_mixedEffects_All_diffGenes.tsv',
                     name = 'Differentially spliced junctions comparison list',
                     parentId = CODE$properties$id)

synSetAnnotations(SIG.SETS_OBJ) <- list(q=0.05, fc=0)
SIG.SETS_OBJ <- synStore(SIG.SETS_OBJ, used = ALL_USED_IDs, 
                         activityName = activityName, executed=thisFile)
```
|  *Results*        |  *SynapseID*                    |
|  -------          |   ---------                     |
|  Log fold change  |  `r paste(FC_OBJ$properties$id,FC_OBJ$properties$versionNumber,sep='.')`                               |
|  Adjusted pvalues |  `r paste(PVAL_OBJ$properties$id,PVAL_OBJ$properties$versionNumber,sep='.')`                           |
|  Differentially spliced junctions   |  `r paste(SIG.SETS_OBJ$properties$id,SIG.SETS_OBJ$properties$versionNumber,sep='.')` |
  
### Executed: Source code
`r paste0('[Source R Markdown](',print(thisFile),')')`