---
title: "Outlier Analysis using eXpress aligned mRNA"
author: "Thanneer Perumal and Kenneth Daily"
date: "03/24/2015"
output: html_document
---
  
```{r libs, echo=FALSE, warning=FALSE, message=FALSE}
## It is assumed your working directory is where this file is
# Load required libraries
library('synapseClient')
library('WGCNA')
library('RColorBrewer')
library('ggplot2')
library('edgeR')
library('fpc')
library('dendextend')

# devtools::source_gist('2866ef5c0aeb64d265ed')
# knit2synapse(file = "./OutlierAnalysis_Express.Rmd", owner = 'syn3382231', wikiName = "Outlier Analysis Express",overwrite=T)
synapseLogin()

# source utility files from ../R/lib folder
# file.sources = list.files('../R/lib',pattern="*.R")
req_files = c('getGeneFilteredGeneExprMatrix.R')
file.sources = sapply(req_files,function(x){return(paste('../R/lib',x,sep='/'))})
tmp = sapply(file.sources,source,.GlobalEnv)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r Input parameters}
# Input Parameters
COUNT_ID = 'syn3354716';
METADATA_ID = 'syn3156503';

# Synapse Specific Parameters
SYNAPSE_STORE = T;
parentId = 'syn3276097';

# Specify factor and continuous covarites pool
FactorCovariates = c('Diffname_short','run','lane','index','Cell_Type_of_Origin',
                     'Tissue_of_Origin', 'Reprogramming_Vector_Type','Reprogramming_Gene_Combination',
                     'Donor_Life_Stage','Originating_Lab','Gender','High_Confidence_Donor_ID')
ContCovariates = c('numberReads','mappedReads')

```

Synapse id of count matrix used for the analysis is `r COUNT_ID` and the synapse id of meta data table used for the analysis is `r METADATA_ID`.

Obtain count matrix and metadata from synapse.

```{r getdata, cache=TRUE}
## Get mRNA counts and metadata table from synapse
# Get count matrix
COUNT_OBJ = synGet(COUNT_ID);
COUNT = read.table(getFileLocation(COUNT_OBJ), sep=',', header=T, row.names=1, check.names=FALSE)

# Get metadata
METADATA_OBJ = synTableQuery(paste('SELECT * FROM',METADATA_ID,sep=' '),verbose = F)
METADATA = METADATA_OBJ@values

ngenes <- dim(COUNT)[1]
nsamples <- dim(COUNT)[2]
```

mRNA data has `r ngenes` genes and `r nsamples` samples 

Preporcess counts & metadata

```{r preprocess data}
# Preprocess metadata
METADATA[METADATA == 'N/A'] = NA

# Arrange count and metadata
RowsPresent = match(colnames(COUNT), METADATA$UID);
METADATA = METADATA[RowsPresent,]
rownames(METADATA) = METADATA$UID

# Preprocess count matrix
COUNT <- na.omit(COUNT) # listwise deletion of missing gene

# Scale count matrix
SCOUNT <- scale(COUNT) # standardize variables for each column (sample)
```

Normalize count data using voom with NULL design (this is performed to stabilise mean-variance relationship)

``` {r normalise count data}
# Get cpm counts (without any filtering)
PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT,MIN_GENE_CPM = 0,MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM = 0)

# Without using any covariates (NULL design to stabilize mean variance trend)
VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=NULL, plot=F)

# Scale expression matrix
SEXPR <- scale(VOOM.GENE_EXPRESSION$E) # standardize variables for each column (sample)
```

Find MDS of scaled mRNA expression

```{r MDS}
# Find MDS to plot
d <- dist(t(SEXPR)) # euclidean distances between the samples
fit <- cmdscale(d,eig=TRUE, k=2) # k is the number of dim
```

Find PCA of scaled mRNA expression

``` {r pca}
# Find principal components to plot
PC <- prcomp(SEXPR)
```

Find outliers using hierarichal clutering

```{r hcluster}
# Get intresting covariates to plot
COVARIATES <- METADATA[,c(FactorCovariates,ContCovariates)]
COVARIATES <- apply(COVARIATES,2,function(x){as.numeric(as.factor(x))})
COVARIATES[is.na(COVARIATES)] <- 0
rownames(COVARIATES) <- rownames(METADATA)

# get clustering for raw counts and normalised expresion matrix
htree <- hclust(dist(t(SEXPR),method="euclidean"),method="average")

# Finding number of clusters using dynamic tree cut package
dynamicMods = cutreeDynamic(dendro = htree, method='tree',pamRespectsDendro = FALSE, minClusterSize = 5);

# Get dendrogram
dend <- as.dendrogram(htree)
```

Plot dendrograms for hierarichal clustering 

``` {r hclustDendro, fig.width=30, fig.height=20}
# Convert covariates to colors
traitcolors <- numbers2colors(COVARIATES,colors=brewer.pal(12,'Paired'),signed=F)
colnames(traitcolors) <- c(FactorCovariates,ContCovariates)

# Plot dendrogram clusters
par(mar = c(30,10,10,2)+0.1)
dend = color_branches(dend,k=max(dynamicMods)+1)
dend = color_labels(dend,k=max(dynamicMods)+1)
plot(dend,ylab='Eucledian distance',main='Normalised mRNA expression with NULL desing');
colored_bars(colors=traitcolors,dend=dend,y_shift = -35)
```

Map hclust data in MDS plot

```{r hclustmds, fig.width=20, fig.height=20}
plotdata <- data.frame(UID=rownames(fit[[1]]), C1=fit[[1]][,1], C2=fit[[1]][,2], cluster=dynamicMods)
plotdata <- merge(plotdata, METADATA, by="UID")

useForShapes <- c("MESO-5", "MESO-15", "MESO-30", "DE", "EB", "SC", "ECTO")
plotdata$plotshape <- ifelse(plotdata$Diffname_short %in% useForShapes,plotdata$Diffname_short, "Other")
plotdata$plotshape <- gsub("-", "", plotdata$plotshape)

potentialOutliers <- unique(c('SC13-054.182.3.7','SC12-040.420.12.19','SC13-043DE.419.12.8','SC12-024ECTO.420.12.12', 'SC13-057iECScramble.219.6.22','SC12-034MESO.219.1.7', 'SC12-039MESO.419.12.27', 'H9.144.7.7',METADATA$UID[!METADATA$pass_qc]))
plotdata$labels <- ifelse(plotdata$UID %in% potentialOutliers, as.character(plotdata$UID),"")

p <- ggplot(plotdata, aes(x=C1, y=C2)) 
p <- p + geom_point(aes(color=factor(cluster),size = pass_qc, shape=plotshape))
p <- p + geom_text(aes(label=labels), size=3, hjust=0) + theme_bw() + scale_size_manual(values = c(5, 2)) + scale_shape_manual(values = c(15,5,16,17,1,18,8,11)) + theme(legend.position="top")
p
```

Map hclust data in PCA plot

```{r hclustpca, fig.width=20,fig.height=20}
# Plot first 2 PCs
plotdata <- data.frame(UID=rownames(PC$rotation), PC1=PC$rotation[,1], PC2=PC$rotation[,2], cluster=dynamicMods)
plotdata <- merge(plotdata, METADATA, by="UID")
plotdata$plotshape <- ifelse(plotdata$Diffname_short %in% c("MESO-5", "MESO-15", "MESO-30", "DE", "EB", "SC", "ECTO"), plotdata$Diffname_short, "Other")
plotdata$plotshape <- gsub("-", "", plotdata$plotshape)
plotdata$outlier <- "no"

plotdata$labels <- ifelse(plotdata$UID %in% potentialOutliers,as.character(plotdata$UID),"")

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=factor(cluster), size=pass_qc, shape=plotshape))
p <- p + theme_bw() + scale_size_manual(values = c(5, 2)) + scale_shape_manual(values = c(15,5,16,17,1,18,8,11)) + theme(legend.position="top") + geom_text(aes(label=labels), size=3, hjust=0)
p
```

Based on hierarichal clustering following observations were made

1. MESO5 tend to cluster with DE samples than MESO15 and 30 which cluster with EB
2. Somatic samples cluster seperately
3. As with cufflinks data, SC13−054.182.3.7, SC12−040.420.12.19, SC13-057iECScramble.219.6.22, H9.133.1.11, SC11-013A.133.5+6.10, SC11-009A.133.2.14, SC11-010A.133.5+6.8, SC11-008A.133.2.13, SC11-012A.133.2.9 samples were observed to be outliers
4. Other than that following 12 samples were identified as outliers SC12-034MESO.219.1.7, SC11-013BEB.144.6.11, SC11-013BEB.149.6.19, SC11-018BEB.144.7.6, SC11-008A.133.2.13, SC11-008BEB.144.6.5,	SC11-010AEB.133.3.19, SC11-017AEB.144.7.2, SC11-017AEB.149.6.20, SC11-017BEB.144.7.4,	SC11-017BEB.149.6.21,	SC11-018AEB.144.7.5, SC11-018AEB.149.6.22, SC11-018BEB.149.6.23, SC13-049DE.219.6.6

Find outliers using kmeans clustering

``` {r kmeansclustering, cahce=T}
# Determine number of clusters automatically using pamk
pam.cluster <- pamk(t(SEXPR), krange=2:10)
```

Map kmeans clustering data in MDS plot

```{r kmeansmds, fig.width=20, fig.height=20}
plotdata <- data.frame(UID=rownames(fit[[1]]), C1=fit[[1]][,1], C2=fit[[1]][,2], 
                       cluster=pam.cluster$pamobject$clustering)
plotdata <- merge(plotdata, METADATA, by="UID")

useForShapes <- c("MESO-5", "MESO-15", "MESO-30", "DE", "EB", "SC", "ECTO")
plotdata$plotshape <- ifelse(plotdata$Diffname_short %in% useForShapes, 
                             plotdata$Diffname_short, "Other")

plotdata$plotshape <- gsub("-", "", plotdata$plotshape)

plotdata$labels <- ifelse(plotdata$UID %in% potentialOutliers, as.character(plotdata$UID),"")

p <- ggplot(plotdata, aes(x=C1, y=C2)) 
p <- p + geom_point(aes(color=factor(cluster),size=pass_qc, shape=plotshape))
p <- p + theme_bw() + scale_size_manual(values = c(5, 2)) + scale_shape_manual(values = c(15,5,16,17,1,18,8,11)) + theme(legend.position="top") + geom_text(aes(label=labels), size=3, hjust=0)
p
```

Map kmeans clustering data in PCA plot

```{r kmeanspca, fig.width=20, fig.height=20}
# Plot first 2 PCs
plotdata <- data.frame(UID=rownames(PC$rotation), PC1=PC$rotation[,1], PC2=PC$rotation[,2], cluster=pam.cluster$pamobject$clustering)
plotdata <- merge(plotdata, METADATA, by="UID")
plotdata$plotshape <- ifelse(plotdata$Diffname_short %in% c("MESO-5", "MESO-15", "MESO-30", "DE", "EB", "SC", "ECTO"), plotdata$Diffname_short, "Other")

plotdata$plotshape <- gsub("-", "", plotdata$plotshape)
plotdata$outlier <- "no"

plotdata$labels <- ifelse(plotdata$UID %in% potentialOutliers,as.character(plotdata$UID),"")

plotdata$outlier <- "no"

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=factor(cluster), size=pass_qc, shape=plotshape))
p <- p + theme_bw() + scale_size_manual(values = c(5, 2)) + scale_shape_manual(values = c(15,5,16,17,1,18,8,11)) + theme(legend.position="top") + geom_text(aes(label=labels), size=3, hjust=0)
p
```

Based on kmeans clustering following observations were made
 
1. MESO5 tend to cluster with DE samples than MESO15 and 30 which cluster with EB
2. Somatic samples cluster seperately
3. As with cufflinks data, SC13−054.182.3.7, SC12−040.420.12.19, SC13-057iECScramble.219.6.22, H9.133.1.11, SC11-013A.133.5+6.10, SC11-009A.133.2.14, SC11-010A.133.5+6.8, SC11-008A.133.2.13, SC11-012A.133.2.9 samples were observed to be outliers
4. Other than that following 12 samples were identified as outliers SC12-034MESO.219.1.7, SC11-013BEB.144.6.11, SC11-013BEB.149.6.19, SC11-018BEB.144.7.6, SC11-008A.133.2.13, SC11-008BEB.144.6.5,  SC11-010AEB.133.3.19, SC11-017AEB.144.7.2, SC11-017AEB.149.6.20, SC11-017BEB.144.7.4,	SC11-017BEB.149.6.21,	SC11-018AEB.144.7.5, SC11-018AEB.149.6.22, SC11-018BEB.149.6.23, SC13-049DE.219.6.6
5. SC12-040 has abnormal karyotype (but not marked exclude and passes qc)
6. SC13-054 did not pass QC
7. SC13-043DE.419.12.8 is a mislabe

```{r storefiles}
Outliers <- c('SC13−054.182.3.7','SC12−040.420.12.19','SC13-057iECScramble.219.6.22','H9.133.1.11','SC11-013A.133.5+6.10','SC11-009A.133.2.14','SC11-010A.133.5+6.8','SC11-008A.133.2.13', 'SC11-012A.133.2.9','SC12-034MESO.219.1.7','SC11-013BEB.144.6.11','SC11-013BEB.149.6.19','SC11-018BEB.144.7.6','SC11-008A.133.2.13','SC11-008BEB.144.6.5','SC11-010AEB.133.3.19','SC11-017AEB.144.7.2','SC11-017AEB.149.6.20','SC11-017BEB.144.7.4','SC11-017BEB.149.6.21','SC11-018AEB.144.7.5','SC11-018AEB.149.6.22', 'SC11-018BEB.149.6.23','SC13-049DE.219.6.6')
write.table(Outliers,file = './Outliers_mRNA.txt',sep='\t',col.names=F,row.names=F,quote=F)

if (SYNAPSE_STORE){
  OUTLIER_CODE <- File('./OutlierAnalysis_Express.Rmd',name = 'Outlier Analysis Express',parentId = parentId)
  OUTLIER_CODE <- synStore(OUTLIER_CODE, used = c(COUNT_OBJ$properties$id,METADATA_OBJ@schema),activityName='Outlier Analysis for Express', executed='https://github.com/th1vairam/pcbc_c4_analysis/blob/cov_anal/code/Rmd/OutlierAnalysis_Express.Rmd')
  OUTLIERS <- File('./Outliers_mRNA.txt',name= 'Outliers Express mRNA',parentId = parentId)
  OUTLIERS <- synStore(OUTLIERS, used = c(COUNT_OBJ$properties$id,METADATA_OBJ@schema),activityName='Outlier Analysis for Express', executed=OUTLIER_CODE)  
}
```