---
title: Regulation of lineage differentiation
author: "Thanneer Perumal"
date: '`r date()`'
output: pdf_document
---
```{r knit2synapse, eval = FALSE, include=FALSE}
library(synapseClient)
library(knitr)
library(knit2synapse)

synapseLogin()

knit2synapse::knitToFolderEntity(file = 'Section3_RegulationOfLineageDifferentiation.Rmd',
                                 parentId = "syn7162983",
                                 entityName = "3 Regulation of Lineage Differentiation")
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file is

# Clear R console screen output
cat("\014")

# Load libraries
library(CovariateAnalysis)
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(ggplot2)
library(RColorBrewer)
library(ComplexHeatmap)
library(matrixStats)

library(fpc)
library(WGCNA)
library(flashClust)

library(synapseClient)
library(knitr)
library(githubr)

library(SNFtool)
library(igraph)
library(Vennerable)

synapseLogin()

library(circlize)
colfunc <- colorRampPalette(c("white", "red"))

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE,
  fig.align = 'left',
  results = 'asis',
  out.width = '8in')
```

```{r synapseStore.params, include=FALSE}
parentId = "syn7162983"
ALL_USED_IDs = c()

thisFileName = 'Section3_RegulationOfLineageDifferentiation.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/pcbc_c4_analysis", 
                    ref="branch", 
                    refName='publication')

thisFile <- getPermlink(repository = thisRepo,
                        repositoryPath=paste0('code/Rmd/', thisFileName))
```
### Data download from synapse
#### Get raw logCPM counts from synapse
```{r getRawCount, include=FALSE}
expr.id = c(mrna = 'syn5011095', mirna = 'syn5014454', methyl = 'syn4487642')
ALL_USED_IDs = c(ALL_USED_IDs, as.character(expr.id))
expr = lapply(expr.id, downloadFile)
```

#### Get covariates from synapse
```{r covars, include=FALSE}
covariates.id = c(mrna = 'syn5011149', mirna = 'syn5014460', methyl = 'syn4487669')
covariates.table.id = c(mrna = 'syn3156503', mirna = 'syn3219876', methyl = 'syn3156828')
ALL_USED_IDs = c(ALL_USED_IDs, as.character(covariates.id), as.character(covariates.table.id))
covariates = lapply(covariates.id, downloadFile)
covariates.table = lapply(covariates.table.id, function(x){
  synTableQuery(paste('select UID,C4_Cell_Line_ID from',x))@values
})

covariates = mapply(function(x,y){
  left_join(x,y)
},covariates, covariates.table)
```

#### Get all possible TFs from synapse
```{r TFs}
TFs = downloadFile('syn6040938')$feature
ALL_USED_IDs = c(ALL_USED_IDs, 'syn6040938')

```

#### Get differential expression results from synapse
```{r thresholds, echo=TRUE}
pval.th <- 0.01

mrna.lfc.th <- log2(2)
mirna.lfc.th <- log2(2)
methyl.changeBeta.th <- 0.35
splicing.changePSI.th <- 0.2
```

```{r diffexp, include=FALSE}
### Download differential expression results from synapse
diffexp.id = c(mrna = 'syn5013690', mirna = 'syn5014584', methyl = 'syn4527629', splicing = 'syn5049321')
ALL_USED_IDs = c(ALL_USED_IDs, as.character(diffexp.id))

# Diff states in a sensible, quasi time-based ordering
orderedDiffState = sort(c("SC", "MESO5", "DE", "MESO15", "MESO30", "ECTO", "EB"))
Comparisons = apply(combn(sort(orderedDiffState), 2), 2, function(x){paste(x, collapse = '_vs_')})

# Get mrna differential expression results from synapse
diffexp = list()
diffexp$mrna = downloadFile(diffexp.id['mrna']) %>%
  dplyr::rename(feature = GeneSymbol) %>%
  tidyr::separate(Comparison, into = c('V1','Comparison','V3'), sep = '\\__') %>%
  dplyr::filter(adj.P.value <= pval.th & abs(logFC) >= mrna.lfc.th & Comparison %in% Comparisons) %>%
  dplyr::select(feature, Comparison, adj.P.value, logFC) 

# Get mirna differential expression results from synapse
diffexp$mirna = downloadFile(diffexp.id['mirna']) %>%
  dplyr::rename(feature = GeneSymbol) %>%
  tidyr::separate(Comparison, into = c('V1','Comparison','V3'), sep = '\\__') %>%
  dplyr::filter(adj.P.value <= pval.th & abs(logFC) >= mirna.lfc.th & Comparison %in% Comparisons) %>%
  dplyr::select(feature, Comparison, adj.P.value, logFC) 

# Get methyl differential expression results from synapse
diffexp$methyl = downloadFile(diffexp.id['methyl']) %>%
  dplyr::select(-one_of('feature')) %>%
  dplyr::rename(feature = methProbeIDs) %>%
  tidyr::separate(Comparison, into = c('V1','Comparison','V3'), sep = '\\__') %>%
  dplyr::filter(adj.P.value <= pval.th & abs(changeBETA) >= methyl.changeBeta.th & Comparison %in% Comparisons) %>%
  dplyr::select(feature, Comparison, adj.P.value, logFC, changeBETA) 

# Get splicing differential expression results from synapse
diffexp$splicing = downloadFile(diffexp.id['splicing']) %>%
  dplyr::select(-one_of('feature')) %>%
  plyr::rename(c('Minor-Isoform' = 'feature')) %>%
  tidyr::separate(Comparison, into = c('V1','Comparison','V3'), sep = '\\__') %>%
  dplyr::filter(adj.P.value <= pval.th & abs(changePSI) >= splicing.changePSI.th & Comparison %in% Comparisons) %>%
  dplyr::select(feature, Comparison, adj.P.value, logFC, changePSI) 

diffexp = rbindlist(diffexp, use.names = T, fill = T, idcol = 'Assay') %>%
  tidyr::separate(Comparison, c('from.state','to.state'), sep = '_vs_')

diffexp = rbindlist(list(diffexp,
                         diffexp %>%
                           dplyr::rename(to.state = from.state, from.state = to.state) %>%
                           dplyr::mutate(
                             logFC = -logFC, changeBETA = -changeBETA, changePSI = -changePSI)),
                    use.names = T, fill =T) %>%
  dplyr::mutate(Direction = logFC/abs(logFC)) %>%
  dplyr::select(-changePSI)
```
Based on the previous analysis here we define four groups of differentiation stages: SC, DE, ECTO and MESO (which is a intersection of mesoderm at day 15 and 30).

#### Get co-expression results from synapse
```{r coexp, include = FALSE}
coexp.id = 'syn5662375'
ALL_USED_IDs = c(ALL_USED_IDs, coexp = coexp.id)
coexpression = downloadFile(coexp.id) %>%
  dplyr::select(feature, target, feature.assay, target.assay, dataSource,contains('SC')) %>%
  filter(feature.assay != 'splicing', target.assay != 'splicing') %>%
  tidyr::gather(comparison, coexpression, -feature, -target, 
                -feature.assay, -target.assay, -dataSource) %>%
  tidyr::separate(comparison, c('from.state', 'to.state'), sep = '_vs_') %>%
  dplyr::mutate(Assay = paste(feature.assay,target.assay, sep = '_')) %>%
  ddply(.(feature.assay), .fun = function(x){
    if(any(unique(x$feature.assay) %in% c('methyl','mirna')))
      return(filter(x, coexpression <= 0))
    else
      return(x)
  })

writeLines('Total number of possible interactions between assays are')
coexpression %>%
  dplyr::select(feature, target, feature.assay, target.assay) %>%
  unique %>%
  group_by(feature.assay, target.assay) %>%
  summarise(count = n()) %>%
  spread(target.assay, count) %>%
  kable

p = ggplot(coexpression, aes(x = Assay, y = coexpression)) + geom_violin()
p = p + theme(axis.text.x = element_text(angle = 45, hjust = 1))
p

writeLines('Out of all possible interactions filter the top 10% of interactions from both the tails of distribution as active interactions')
coexpression = coexpression %>%
  ddply(.(Assay), .fun = function(x){
    cut_off = quantile(abs(x$coexpression), 0.9, na.rm = T)
    filter(x, abs(coexpression) >= cut_off)
  })
total.interactions = coexpression %>%
  dplyr::select(feature, target, feature.assay, target.assay) %>%
  unique %>%
  group_by(feature.assay, target.assay) %>%
  summarise(count = n()) 

total.interactions %>%
  spread(target.assay, count) %>%
  kable

# Flip states and append interaction list
coexpression = rbindlist(list(coexpression,
                              coexpression %>% 
                                dplyr::rename(from.state = to.state, to.state = from.state)),
                         use.names = T, fill = T) %>%
  filter(from.state == 'SC')
```

### Regulatory network of lineage differentiation
Relative control of TFs vs mirna vs mrna regulations in SC differentiation
```{r relative.control}
all.int = diffexp %>%
  dplyr::rename(target = feature, target.assay = Assay, target.lfc = logFC, target.fdr = adj.P.value,
                target.beta = changeBETA, target.direction = Direction) %>%
  filter(from.state == 'SC', target.assay != 'splicing') %>%
  left_join(coexpression) %>%
  left_join( diffexp %>%
               dplyr::rename(feature.assay = Assay, feature.lfc = logFC, feature.fdr = adj.P.value, 
                             feature.beta = changeBETA, feature.direction = Direction) %>%
               filter(from.state == 'SC', feature.assay != 'splicing')) %>%
  filter(!(target.assay %in% c('mirna','methyl') & is.na(feature.lfc)))

all.int$feature.assay[all.int$feature %in% TFs] = 'TF'
all.int$feature.assay[is.na(all.int$feature.assay)] = 'NONE'
all.int$target.assay[is.na(all.int$target.assay)] = 'NONE'
# all.int$target.assay[all.int$target %in% TFs] = 'TF'
all.int$Assay = paste(all.int$feature.assay, all.int$target.assay, sep = '_')
all.int = filter(all.int, ! (Assay %in% c('mrna_mrna', 'mrna_TF')))

# Combine MESO15 and 30 together
all.int.MESO  = filter(all.int, to.state %in% c('MESO15','MESO30'))
all.int.notMESO  = filter(all.int, to.state %in% c('DE','ECTO','EB','MESO5'))

all.int.MESO = all.int.MESO %>%
  ddply(.(feature, target), .fun = function(x){
    if(dim(x)[1] == 2)
      x = top_n(x,1,abs(coexpression),abs(target.lfc))
    else
      return(NULL)
  })
all.int.MESO$to.state = 'MESO'
all.int = rbindlist(list(all.int.notMESO, all.int.MESO), fill = T, use.names = T)

relative.control = all.int %>%
  ddply(.(to.state, target.assay, feature.assay, target.direction), .fun = function(x){
    count = data.frame(count = x %>%
                         dplyr::select(target) %>%
                         unlist %>% unique %>% length)
    })

total.diffexp = group_by(all.int, to.state, target.assay) %>%
  summarise(count.all = length(unique(target)))

tmp = full_join(relative.control, total.diffexp) %>%
  dplyr::mutate(percent = 100*(count/count.all))

p = tmp %>% 
  filter(target.assay == 'mrna') %>%
  ggplot(aes(x = feature.assay, y = percent, fill = to.state))
p = p + geom_bar(stat = 'identity', position = 'dodge') + ggtitle('Relative regulation of mrna')
p

p = tmp %>% 
  filter(target.assay == 'mrna', to.state %in% c('DE','ECTO','MESO')) %>%
  ggplot(aes(x = feature.assay, y = percent, fill = to.state))
p = p + geom_bar(stat = 'identity', position = 'dodge') + ggtitle('Relative regulation of mrna')
p

p = p + facet_grid(.~target.direction)
p
```

### Store intermediate results in synapse
```{r synapse.store}
# Populate wiki with results
CODE <- Folder(name = '3 Regulation of Lineage Differentiation',parentId = 'syn7162983')
CODE <- synStore(CODE)

# Store active interactions to synapse
all.int %>%
  write.table(file = 'ActiveInteractions.tsv', sep = '\t', quote=F, row.names = F)
obj = File('ActiveInteractions.tsv', 
           name = 'Active Interactions',
           parentId = CODE$properties$id)
obj = synStore(obj, 
               activityName = 'Network Identification', 
               executed = thisFile, 
               used = ALL_USED_IDs)
```